---
title: "Bayesian Updating (Scalar): Fusing Priors with Observed Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bayesian Updating (Scalar): Fusing Priors with Observed Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6, fig.height = 4)
```

## Overview

The other vignettes in this package (`getting-started-monte-carlo.Rmd`, `raster-fusion-ssurgo-solus.Rmd`)
walk through top-level *pipeline* functions end to end. This vignette instead zooms in on a single
thematic group of low-level, pure-math functions - `R/bayesian-updating.R` - and works through each one
individually: what it computes, what each parameter controls, and how varying that parameter changes the
result. It is the finer-grained, "read the primitives" companion to `raster-fusion-ssurgo-solus.Rmd`,
which exercises this exact math implicitly (raster-native, cell by cell) inside its fusion pipeline but
never isolates it. See `soilSIM/docs/08_bayesian_updating.md` for the full function-level reference this
vignette is based on.

`bayesian-updating.R` fuses a **prior** belief about a soil property (e.g. an SSURGO-derived estimate)
with a **likelihood** - newly observed data (e.g. lab or field measurements) - into a **posterior**
belief that combines both, weighted by how confident (precise) each side is. All examples below use
small, clearly-labeled synthetic prior/likelihood values - no network access or real SSURGO/SOLUS data
is needed for this vignette.

Steps:

1. `bayes_update_normal_normal()` - exact closed-form Normal-Normal fusion, worked example plus a
   sensitivity sweep over relative precision.
2. `normal_to_lognormal_params()` / `lognormal_to_normal_params()` - fusing a right-skewed, strictly
   positive property (organic matter) safely, in log-space.
3. `fuse_beta()` / `fuse_gamma()` - closed-form same-family fusion for bounded (Beta) and positive-skewed
   (Gamma) properties, plus their moment-conversion helpers.
4. `fuse_bivariate_normal()` - the 2D generalization, illustrated on a joint clay/sand/silt (ILR) example.
5. The dispatchers `bayes_fuse()`, `bayesian_update()`, and `fuse_property()` - how they route to the
   right method.

```{r setup}
library(soilSIM)
library(ggplot2)
```

## Step 1: `bayes_update_normal_normal()` - exact Normal-Normal fusion

**Purpose**: fuse a Normal prior `N(prior_mu, prior_sigma)` with a Normal likelihood
`N(lik_mu, lik_sigma)` into a Normal posterior, via *precision-weighted averaging*. Precision is the
inverse variance (`1 / sigma^2`) - a tighter (smaller-sigma) distribution has higher precision, i.e. it
represents a more confident belief.

**Parameters**:

- `prior_mu`, `prior_sigma` - the prior's mean and standard deviation.
- `lik_mu`, `lik_sigma` - the likelihood's mean and standard deviation, in the same units.

All four arguments are plain numeric scalars or equal-length vectors (the function is fully vectorized,
which is what lets `raster-fusion.R` reuse it unchanged across every cell of a `SpatRaster`).

### A worked example

A hypothetical SSURGO-derived prior belief about clay content (%) for some map unit, and a hypothetical
lab-observed likelihood from a handful of new samples at the same location:

```{r}
prior_clay <- list(mu = 22, sigma = 6)   # SSURGO-derived prior belief
lik_clay   <- list(mu = 30, sigma = 3)   # hypothetical lab observation, more precise than the prior

posterior_clay <- bayes_update_normal_normal(
  prior_mu = prior_clay$mu, prior_sigma = prior_clay$sigma,
  lik_mu = lik_clay$mu, lik_sigma = lik_clay$sigma
)
posterior_clay
```

Two documented invariants, checked here empirically rather than just asserted:

```{r}
# Fusion only sharpens - the posterior can never be more uncertain than either input
posterior_clay$sigma <= min(prior_clay$sigma, lik_clay$sigma)

# The posterior mean is pulled between the two inputs, never outside their range
posterior_clay$mu >= min(prior_clay$mu, lik_clay$mu) && posterior_clay$mu <= max(prior_clay$mu, lik_clay$mu)
```

And the function is symmetric under swapping which side is called "prior" and which is "likelihood" -
the math doesn't care about the labels, only the (mu, sigma) pairs:

```{r}
swapped <- bayes_update_normal_normal(
  prior_mu = lik_clay$mu, prior_sigma = lik_clay$sigma,
  lik_mu = prior_clay$mu, lik_sigma = prior_clay$sigma
)
identical(posterior_clay, swapped)
```

Plotting all three distributions makes the fusion concrete: the posterior (green) is narrower than
*both* the prior (purple) and the likelihood (yellow), and sits closer to the likelihood because the
likelihood was specified as more precise (smaller sigma) here:

```{r, fig.height = 4}
normal_pdf_df <- function(mu, sigma, label, x) {
  data.frame(x = x, density = dnorm(x, mu, sigma), source = label)
}

x_grid <- seq(0, 45, length.out = 400)
dist_df <- rbind(
  normal_pdf_df(prior_clay$mu, prior_clay$sigma, "Prior", x_grid),
  normal_pdf_df(lik_clay$mu, lik_clay$sigma, "Likelihood", x_grid),
  normal_pdf_df(posterior_clay$mu, posterior_clay$sigma, "Posterior", x_grid)
)
dist_df$source <- factor(dist_df$source, levels = c("Prior", "Likelihood", "Posterior"))

ggplot(dist_df, aes(x = x, y = density, color = source, fill = source)) +
  geom_area(alpha = 0.25, position = "identity") +
  geom_line(linewidth = 1) +
  scale_color_viridis_d(name = NULL) +
  scale_fill_viridis_d(name = NULL) +
  theme_minimal() +
  labs(
    title = "Normal-Normal Bayesian fusion: clay content",
    subtitle = "Posterior is narrower than both inputs, and pulled toward the more precise (likelihood) side",
    x = "Clay content (%)", y = "Density"
  )
```

### Sensitivity sweep: relative precision drives where the posterior lands

The core intuition of Bayesian fusion is that the posterior mean is not a simple average of the two
inputs - it's a *precision-weighted* average, so whichever side is more certain (smaller sigma) pulls
the posterior further toward itself. We hold `prior_mu`/`prior_sigma` fixed and vary only `lik_sigma`
across three cases - likelihood much more precise, comparably precise, and much less precise than the
prior - while holding `lik_mu` fixed too, so any movement in the posterior mean is attributable purely
to the changing relative precision:

```{r}
prior_mu <- 22; prior_sigma <- 6
lik_mu <- 30

sweep_cases <- data.frame(
  case = c("Likelihood much more precise", "Comparable precision", "Likelihood much less precise"),
  lik_sigma = c(1, 6, 18)
)

sweep_results <- do.call(rbind, lapply(seq_len(nrow(sweep_cases)), function(i) {
  post <- bayes_update_normal_normal(prior_mu, prior_sigma, lik_mu, sweep_cases$lik_sigma[i])
  data.frame(
    case = sweep_cases$case[i],
    lik_sigma = sweep_cases$lik_sigma[i],
    posterior_mu = post$mu,
    posterior_sigma = post$sigma
  )
}))
sweep_results
```

As `lik_sigma` shrinks (likelihood becomes more precise, first row), `posterior_mu` moves closer to
`lik_mu` (30); as `lik_sigma` grows (likelihood becomes less precise, last row), `posterior_mu` stays
closer to `prior_mu` (22). This is the fusion "tug of war" made visible:

```{r, fig.height = 4}
ggplot(sweep_results, aes(x = reorder(case, lik_sigma), y = posterior_mu)) +
  geom_hline(yintercept = prior_mu, linetype = "dashed", color = viridisLite::viridis(1, begin = 0.15)) +
  geom_hline(yintercept = lik_mu, linetype = "dashed", color = viridisLite::viridis(1, begin = 0.75)) +
  geom_point(aes(color = case), size = 4) +
  geom_errorbar(aes(ymin = posterior_mu - posterior_sigma, ymax = posterior_mu + posterior_sigma,
                     color = case), width = 0.15) +
  scale_color_viridis_d(guide = "none") +
  theme_minimal() +
  labs(
    title = "Posterior mean shifts toward whichever side is more precise",
    subtitle = "Dashed lines mark prior_mu (22, lower) and lik_mu (30, upper); prior_sigma fixed at 6",
    x = NULL, y = "Posterior mean (+/- posterior sd)"
  )
```

## Step 2: fusing a right-skewed property safely - `normal_to_lognormal_params()` / `lognormal_to_normal_params()`

`bayes_update_normal_normal()` assumes both sides are Normal. That's a problem for strictly-positive,
right-skewed properties like organic matter (OM) or CEC: if a property's sigma is large relative to its
mean, treating it as Normal and fusing directly can place real posterior probability mass *below zero*,
which is physically impossible for a percentage or concentration.

`normal_to_lognormal_params(mu, sigma)` moment-matches a raw-space Lognormal's mean/sd onto the `mu`/
`sigma` of its underlying (log-space) Normal, so `bayes_update_normal_normal()` can be reused unchanged
in log-space; `lognormal_to_normal_params(mu_log, sigma_log)` is its exact algebraic inverse, converting
the fused log-space posterior back to raw-space mean/sd.

A hypothetical prior and likelihood for organic matter (%), a property that is right-skewed with a
sigma large enough, relative to its mean, to be a real concern:

```{r}
prior_om <- list(mu = 3.0, sigma = 2.0)   # SSURGO-derived prior; sigma is 2/3 of the mean
lik_om   <- list(mu = 5.5, sigma = 1.0)   # hypothetical lab observation
```

Fusing these directly as if Normal risks a posterior mean minus a couple of sigma dropping below zero;
more importantly, the *shape* of a raw Normal fit ignores the right skew altogether. Converting to
log-space first:

```{r}
prior_om_log <- normal_to_lognormal_params(prior_om$mu, prior_om$sigma)
lik_om_log <- normal_to_lognormal_params(lik_om$mu, lik_om$sigma)
prior_om_log
lik_om_log
```

Fusing in log-space (still an exact Normal-Normal update, just on the log-transformed parameters), then
converting the posterior back to raw space:

```{r}
posterior_om_log <- bayes_update_normal_normal(
  prior_om_log$mu, prior_om_log$sigma, lik_om_log$mu, lik_om_log$sigma
)
posterior_om <- lognormal_to_normal_params(posterior_om_log$mu, posterior_om_log$sigma)
posterior_om
```

`lognormal_to_normal_params()` is the exact inverse of `normal_to_lognormal_params()` - round-tripping
raw-space parameters through both functions returns the original values, up to floating-point precision:

```{r}
roundtrip <- lognormal_to_normal_params(prior_om_log$mu, prior_om_log$sigma)
all.equal(roundtrip, prior_om)
```

Plotting the raw-space prior, likelihood, and posterior as Lognormal densities (never negative, right
skewed, exactly what a percentage should look like) versus what a naive raw-space Normal fusion would
have produced:

```{r, fig.height = 4}
x_grid_om <- seq(0.001, 12, length.out = 400)

lognormal_pdf <- function(mu_log, sigma_log, label, x) {
  data.frame(x = x, density = dlnorm(x, meanlog = mu_log, sdlog = sigma_log), source = label)
}

om_dist_df <- rbind(
  lognormal_pdf(prior_om_log$mu, prior_om_log$sigma, "Prior (Lognormal)", x_grid_om),
  lognormal_pdf(lik_om_log$mu, lik_om_log$sigma, "Likelihood (Lognormal)", x_grid_om),
  lognormal_pdf(posterior_om_log$mu, posterior_om_log$sigma, "Posterior (Lognormal)", x_grid_om)
)
om_dist_df$source <- factor(om_dist_df$source,
  levels = c("Prior (Lognormal)", "Likelihood (Lognormal)", "Posterior (Lognormal)"))

naive_posterior_om <- bayes_update_normal_normal(prior_om$mu, prior_om$sigma, lik_om$mu, lik_om$sigma)

ggplot(om_dist_df, aes(x = x, y = density, color = source, fill = source)) +
  geom_area(alpha = 0.25, position = "identity") +
  geom_line(linewidth = 1) +
  geom_vline(xintercept = naive_posterior_om$mu - 2 * naive_posterior_om$sigma,
             linetype = "dotted", color = "grey40") +
  annotate("text", x = naive_posterior_om$mu - 2 * naive_posterior_om$sigma, y = 0, hjust = -0.05,
           vjust = -0.5, size = 3, color = "grey40",
           label = "naive raw-Normal\nposterior mean - 2sd") +
  scale_color_viridis_d(name = NULL) +
  scale_fill_viridis_d(name = NULL) +
  theme_minimal() +
  labs(
    title = "Log-space fusion keeps organic matter strictly positive",
    subtitle = "The naive raw-space Normal posterior's mean-2sd point (dotted line) falls near/below zero",
    x = "Organic matter (%)", y = "Density"
  )
```

```{r}
# The naive raw-space Normal fusion's implied lower tail:
naive_posterior_om$mu - 2 * naive_posterior_om$sigma
```

## Step 3: same-family closed-form fusion - `fuse_beta()` / `fuse_gamma()`

Some properties are naturally bounded (a fraction/percentage in `[0, 1]`, e.g. saturation) and fit
better as a **Beta** distribution than a Normal; others are strictly positive with a long right tail and
fit better as a **Gamma** distribution. Both families have their own exact closed-form conjugate fusion
rule, analogous to Normal precision-addition but on that family's own shape parameters.

### `fuse_beta()`

**Purpose**: fuse two Beta beliefs, `Beta(prior_alpha, prior_beta)` and `Beta(lik_alpha, lik_beta)`, via
straight addition of shape parameters minus one: `Beta(a1,b1) x Beta(a2,b2) -> Beta(a1+a2-1, b1+b2-1)`.

**Parameters**: `prior_alpha`, `prior_beta`, `lik_alpha`, `lik_beta` - each side's Beta shape parameters
(numeric scalar or vector).

**Returns**: `list(alpha=, beta=, feasible=)` - `feasible` flags whether the fused shape parameters
stayed positive (`prior_alpha + lik_alpha > 1` and `prior_beta + lik_beta > 1`), a requirement for the
result to remain a valid distribution.

A hypothetical prior and likelihood belief about water-saturation fraction, both expressed as Beta
parameters (moment-matched via `moments_to_beta()` from an assumed mean/variance, so the shapes are
concrete and interpretable):

```{r}
prior_sat_moments <- list(mean = 0.35, var = 0.02^2)
lik_sat_moments <- list(mean = 0.50, var = 0.01^2)

prior_sat_beta <- moments_to_beta(prior_sat_moments$mean, prior_sat_moments$var)
lik_sat_beta <- moments_to_beta(lik_sat_moments$mean, lik_sat_moments$var)
prior_sat_beta
lik_sat_beta

posterior_sat_beta <- fuse_beta(
  prior_sat_beta$alpha, prior_sat_beta$beta, lik_sat_beta$alpha, lik_sat_beta$beta
)
posterior_sat_beta
```

`beta_to_moments()` converts the fused shape parameters back to an interpretable mean/variance, and is
also the function used (on the *inputs*) if `feasible` had come back `FALSE`, to fall back to fusing in
Normal-moment space instead:

```{r}
beta_to_moments(posterior_sat_beta$alpha, posterior_sat_beta$beta)
```

```{r, fig.height = 4}
beta_pdf <- function(alpha, beta, label, x) {
  data.frame(x = x, density = dbeta(x, alpha, beta), source = label)
}
x_grid_sat <- seq(0, 1, length.out = 400)
sat_dist_df <- rbind(
  beta_pdf(prior_sat_beta$alpha, prior_sat_beta$beta, "Prior", x_grid_sat),
  beta_pdf(lik_sat_beta$alpha, lik_sat_beta$beta, "Likelihood", x_grid_sat),
  beta_pdf(posterior_sat_beta$alpha, posterior_sat_beta$beta, "Posterior", x_grid_sat)
)
sat_dist_df$source <- factor(sat_dist_df$source, levels = c("Prior", "Likelihood", "Posterior"))

ggplot(sat_dist_df, aes(x = x, y = density, color = source, fill = source)) +
  geom_area(alpha = 0.25, position = "identity") +
  geom_line(linewidth = 1) +
  scale_color_viridis_d(name = NULL) +
  scale_fill_viridis_d(name = NULL) +
  theme_minimal() +
  labs(
    title = "Beta-Beta conjugate fusion: water-saturation fraction",
    x = "Saturation fraction", y = "Density"
  )
```

### `fuse_gamma()`

**Purpose**: fuse two Gamma beliefs, `Gamma(prior_shape, prior_rate)` and `Gamma(lik_shape, lik_rate)`
(shape/rate parameterization), via `Gamma(k1,r1) x Gamma(k2,r2) -> Gamma(k1+k2-1, r1+r2)` - rates add
outright (like Normal precisions), shapes add minus one (like Beta's alpha/beta).

**Parameters**: `prior_shape`, `prior_rate`, `lik_shape`, `lik_rate`.

**Returns**: `list(shape=, rate=, feasible=)`, `feasible` requiring `prior_shape + lik_shape > 1`.

A hypothetical prior/likelihood for saturated hydraulic conductivity (strictly positive, long right
tail), again built via the Gamma method-of-moments helper `moments_to_gamma()`:

```{r}
prior_ksat_moments <- list(mean = 15, var = 6^2)
lik_ksat_moments <- list(mean = 8, var = 2^2)

prior_ksat_gamma <- moments_to_gamma(prior_ksat_moments$mean, prior_ksat_moments$var)
lik_ksat_gamma <- moments_to_gamma(lik_ksat_moments$mean, lik_ksat_moments$var)
prior_ksat_gamma
lik_ksat_gamma

posterior_ksat_gamma <- fuse_gamma(
  prior_ksat_gamma$shape, prior_ksat_gamma$rate, lik_ksat_gamma$shape, lik_ksat_gamma$rate
)
posterior_ksat_gamma
gamma_to_moments(posterior_ksat_gamma$shape, posterior_ksat_gamma$rate)
```

```{r, fig.height = 4}
gamma_pdf <- function(shape, rate, label, x) {
  data.frame(x = x, density = dgamma(x, shape = shape, rate = rate), source = label)
}
x_grid_ksat <- seq(0.01, 40, length.out = 400)
ksat_dist_df <- rbind(
  gamma_pdf(prior_ksat_gamma$shape, prior_ksat_gamma$rate, "Prior", x_grid_ksat),
  gamma_pdf(lik_ksat_gamma$shape, lik_ksat_gamma$rate, "Likelihood", x_grid_ksat),
  gamma_pdf(posterior_ksat_gamma$shape, posterior_ksat_gamma$rate, "Posterior", x_grid_ksat)
)
ksat_dist_df$source <- factor(ksat_dist_df$source, levels = c("Prior", "Likelihood", "Posterior"))

ggplot(ksat_dist_df, aes(x = x, y = density, color = source, fill = source)) +
  geom_area(alpha = 0.25, position = "identity") +
  geom_line(linewidth = 1) +
  scale_color_viridis_d(name = NULL) +
  scale_fill_viridis_d(name = NULL) +
  theme_minimal() +
  labs(
    title = "Gamma-Gamma conjugate fusion: saturated hydraulic conductivity",
    x = "Ksat (cm/hr)", y = "Density"
  )
```

Both `fuse_beta()` and `fuse_gamma()` are only valid when the fused shape parameters stay positive - if
`feasible` comes back `FALSE` for some cell/case, the documented fallback is to re-express both sides as
Normal moments (`beta_to_moments()`/`gamma_to_moments()`), fuse with `bayes_update_normal_normal()`, and
convert the result back (`moments_to_beta()`/`moments_to_gamma()`). This fallback is not automatic - a
caller hitting `feasible == FALSE` must invoke it explicitly.

## Step 4: joint 2D fusion - `fuse_bivariate_normal()`

Clay, sand, and silt must sum to 100 - fusing them independently (three separate `fuse_beta()` calls)
measurably breaks that constraint. `fuse_bivariate_normal()` is the direct multivariate generalization of
`bayes_update_normal_normal()`: precision *matrices* add instead of scalar precisions, which is what lets
soilSIM fuse clay/sand/silt jointly in 2D isometric-log-ratio (ILR) space and get a fused composition
that sums to 100 by construction.

**Parameters**: `mu1`, `mu2` - length-2 mean vectors; `Sigma1`, `Sigma2` - 2x2 covariance matrices.

**Returns**: `list(mu = fused mean vector, Sigma = fused covariance matrix)`.

A small, illustrative example directly in ILR-coordinate space (not full clay/sand/silt triplets - that
full workflow is `fuse_texture_group_from_triplets()`, which wraps this function together with
`distributions.R`'s `estimate_ilr_moments_mc()`/`sample_ilr_posterior()`; see
`soilSIM/docs/08_bayesian_updating.md` for that whole chain). Here we fuse two independent bivariate
Normal beliefs about a 2D ILR coordinate pair directly:

```{r}
mu1 <- c(0.10, -0.30)                      # hypothetical prior ILR-space belief
Sigma1 <- matrix(c(0.05^2, 0, 0, 0.04^2), nrow = 2)

mu2 <- c(0.25, -0.15)                      # hypothetical likelihood ILR-space belief
Sigma2 <- matrix(c(0.02^2, 0, 0, 0.02^2), nrow = 2)

fused_ilr <- fuse_bivariate_normal(mu1, Sigma1, mu2, Sigma2)
fused_ilr
```

As in the scalar case, fusion sharpens: each posterior variance (the diagonal of `Sigma`) is no larger
than the corresponding diagonal entries of either input covariance matrix, and the posterior mean sits
between the two inputs, pulled toward whichever side had the smaller (more precise) variances -
`Sigma2` here, so `fused_ilr$mu` sits closer to `mu2`:

```{r}
diag(fused_ilr$Sigma) <= pmin(diag(Sigma1), diag(Sigma2))
```

`fuse_texture_group_from_triplets(prior_triplets, lik_triplets, z_prior, z_lik, n_mc, n_samples, total)`
is the full clay/sand/silt workflow built on top of this: it estimates each side's ILR-space `(mu,
Sigma)` from low/representative/high triplets via `estimate_ilr_moments_mc()`, fuses them with
`fuse_bivariate_normal()` exactly as above, then draws posterior clay/sand/silt samples via
`sample_ilr_posterior()` - guaranteed, by construction, to sum to 100. `raster-fusion-ssurgo-solus.Rmd`'s
"Step 2: Fuse a compositional group" demonstrates this full chain on real SSURGO/SOLUS100 rasters.

## Step 5: the dispatchers - `bayes_fuse()`, `bayesian_update()`, `fuse_property()`

Rather than calling `bayes_update_normal_normal()`/`fuse_beta()`/`fuse_gamma()` directly, code often goes
through one of three dispatchers that route to the right method automatically.

### `bayes_fuse()` - same-family dispatcher

**Purpose**: pick the right closed-form `fuse_*()` function based on a `family` argument, given both
sides' parameters as named lists.

**Parameters**: `prior_params`, `lik_params` - named lists matching `family` (`list(mu=,sigma=)` for
`"normal"`, `list(alpha=,beta=)` for `"beta"`, `list(shape=,rate=)` for `"gamma"`); `family` - one of
`"normal"`, `"beta"`, `"gamma"` (via `match.arg()`).

```{r}
bayes_fuse(prior_clay, lik_clay, family = "normal")
bayes_fuse(list(alpha = prior_sat_beta$alpha, beta = prior_sat_beta$beta),
           list(alpha = lik_sat_beta$alpha, beta = lik_sat_beta$beta),
           family = "beta")
```

It is a plain `switch()` on `family` with no logic beyond argument routing - both sides must already be
fit in the same family; there's no closed form for fusing mismatched families.

### `bayesian_update()` - fully general grid-KDE fusion

**Purpose**: when no distributional family fits cleanly, or the two sides are in different families,
`bayesian_update()` fuses raw samples (not parameters) on a shared value grid via kernel density
estimation - fully general, at higher computational cost than the closed-form tiers.

**Parameters**: `prior_distribution`, `likelihood_distribution` - numeric vectors of raw samples;
`grid_range` - optional length-2 min/max (default `NULL`, derived from the combined data range padded by
1); `grid_resolution` - grid step size (default `0.01`); `n` - number of posterior samples to draw
(default `1000`).

```{r}
set.seed(42)
prior_samples <- rnorm(500, mean = prior_clay$mu, sd = prior_clay$sigma)
lik_samples <- rnorm(30, mean = lik_clay$mu, sd = lik_clay$sigma)

posterior_samples <- bayesian_update(prior_samples, lik_samples, grid_resolution = 0.01, n = 1000)
mean(posterior_samples)
sd(posterior_samples)
```

For comparison, the closed-form result computed the same way in Step 1 for this exact prior/likelihood
pair:

```{r}
posterior_clay$mu
posterior_clay$sigma
```

The sample-based mean/sd approximate the closed-form values, as expected - `bayesian_update()` is the
fully general fallback for cases the closed-form tiers can't handle (arbitrary shapes, mismatched
families), not a routine substitute for them.

### `fuse_property()` - top-level entry point, dispatching by input shape

**Purpose**: a single entry point that inspects the *shape* of `prior`/`likelihood` (raw-sample vectors
vs. named parameter lists) rather than requiring a caller-supplied method flag, and routes accordingly.

**Parameters**: `prior`, `likelihood` - both raw-sample vectors, or both named parameter lists matching
`family`; `family` - required for the parameter-list route; `bounds` - unused, kept for interface
symmetry; `method` - optional assertion (`"general"` or `"closed_form"`); errors if it doesn't match
what the input shape implies. `n_samples`, `grid_resolution` - passed through to `bayesian_update()`.

```{r}
# Vector inputs -> routes to bayesian_update() (the general route)
fuse_property(prior_samples, lik_samples, n_samples = 500) |> (\(s) c(mean = mean(s), sd = sd(s)))()

# Parameter-list inputs -> routes to bayes_fuse() (the closed-form route)
fuse_property(prior_clay, lik_clay, family = "normal")
```

The two routes deliberately return different shapes (a sample vector vs. a parameter list) - not
unified into a fake-common contract, since the caller already knows which shape it passed in and
therefore which shape it gets back. Passing a `method` that contradicts the input shape is an error
rather than a silent override:

```{r, error = TRUE}
fuse_property(prior_clay, lik_clay, family = "normal", method = "general")
```

## Where this fits in soilSIM

`bayesian-updating.R` is a standalone toolkit, intentionally not called anywhere in `monte-carlo.R` -
these are independently-tested building blocks for a future "update an SSURGO-derived prior against
observed/field data" feature. `R/raster-fusion.R` is the one place these exact functions are already
assembled into a complete pipeline, applying them unchanged to `SpatRaster` cell values (via `terra`'s
operator overloading) rather than plain scalars/vectors - see `raster-fusion-ssurgo-solus.Rmd` for that
pipeline in action, and `soilSIM/docs/08_bayesian_updating.md` /
`soilSIM/docs/09_multi_source_raster_fusion_pipeline.md` for the full function-level references behind
both.
