---
title: "Distribution Fitting & Correlations: A Function-by-Function Tour"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Distribution Fitting & Correlations: A Function-by-Function Tour}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6, fig.height = 4)
```

## Overview

soilSIM's other vignettes (e.g. "Getting Started: Tabular Monte Carlo Simulation from SSURGO")
walk through top-level orchestrator functions end to end on real data. This vignette instead zooms
in on one thematic group of *individual* functions - soilSIM's generic, data-source-agnostic
statistical core - and shows what each one does and how its parameters change its behavior. Nothing
here knows or cares whether a percentile triplet came from SSURGO, SOLUS, or a KSSL lab table; these
functions take plain numeric percentiles and physical bounds and hand back fitted-distribution
objects, quantile functions, ILR-transformed compositional coordinates, and repaired
positive-definite correlation matrices. See `soilSIM/docs/03_distribution_fitting_correlations.md`
for the full function-level reference.

All data in this vignette is small and synthetic - hand-constructed numbers, not a real download -
so that each parameter's effect is easy to isolate and see.

1. Fit the same percentile triplet with five different distribution families and compare their
   shapes directly (`fit_normal_triplet()`, Beta fitting, `quantile_linear_cdf()`,
   `quantile_triangular()`, metalog fitting)
2. `resolve_property_family()`'s automatic family selection from skew, and how `skew_threshold`
   changes the decision
3. `fit_percentile_triplet()` as the single master dispatcher tying the families together
4. `fit_beta_mle_newton()`'s `n_iter` parameter - watching the Newton-Raphson fit actually converge
5. ILR (isometric log-ratio) compositional transforms: `ilr_forward()` / `ilr_inverse()`
6. Correlation-matrix repair: `ensure_positive_definite_matrix()` and `validate_correlation_matrix()`

```{r setup}
library(soilSIM)
library(ggplot2)
```

## Step 1: Fitting the same triplet with five different families

A single low/representative/high (l/r/h) percentile triplet, at probabilities 0.05/0.5/0.95, stands
in for a clay-content-like property throughout this section:

```{r}
l <- 15; r <- 25; h <- 40   # clay%, at p = 0.05 / 0.50 / 0.95
bounds <- c(0, 100)          # physical bounds for the bounded families
```

Note this triplet is right-skewed (the high-side spread, `(h - r)` = 15, is wider than the low-side
spread, `(r - l)` = 10) - that skew is what later drives `resolve_property_family()`'s automatic
"lognormal" choice in Step 2.

### Normal: `fit_normal_triplet()` / `quantile_normal()`

`fit_normal_triplet(p_lo_val, p50_val, p_hi_val, p_lo, p_hi)` is an exact closed-form fit - no
iteration. `mean` is just the median value (`p50_val`); `sd` is solved directly from the two-point
spread `(p_hi_val - p_lo_val) / (qnorm(p_hi) - qnorm(p_lo))`, an algebraic consequence of the
Normal's linear quantile function. `quantile_normal(fit, q)` then evaluates `mean + sd * qnorm(q)`
at any vector of probabilities `q`.

```{r}
normal_fit <- fit_normal_triplet(p_lo_val = l, p50_val = r, p_hi_val = h, p_lo = 0.05, p_hi = 0.95)
normal_fit
quantile_normal(normal_fit, q = c(0.05, 0.5, 0.95))  # round-trips back to l, r, h
```

### Beta: `fit_beta_mle_newton()` / `quantile_beta()`

Beta fitting rescales the triplet onto `[0,1]` using `bounds`, then fits `shape1`/`shape2` by
Newton-Raphson maximum likelihood (seeded from the closed-form method-of-moments fit,
`fit_beta_mom()`). `quantile_beta(fit, q)` evaluates `qbeta()` on `[0,1]`; rescaling back to raw
units (`fit_percentile_triplet()`/`quantile_from_fit()` handle that step, shown in Step 3 below) is
not part of `quantile_beta()` itself.

```{r}
beta_fit <- fit_beta_mle_newton(values = c(l, r, h), bounds = bounds)
beta_fit
# quantile_beta() operates on [0,1]; rescale manually here for a direct comparison
bounds[1] + quantile_beta(beta_fit, q = c(0.05, 0.5, 0.95)) * diff(bounds)
```

### Piecewise-linear: `quantile_linear_cdf()`

`quantile_linear_cdf(probs, values, q)` builds a piecewise-linear inverse-CDF through the supplied
knots via `approxfun(method = "linear", rule = 2)` - exact at the knots, with `rule = 2` clamping
(not extrapolating) beyond the outermost knot.

```{r}
quantile_linear_cdf(probs = c(0.05, 0.5, 0.95), values = c(l, r, h), q = c(0.05, 0.5, 0.95))
quantile_linear_cdf(probs = c(0.05, 0.5, 0.95), values = c(l, r, h), q = c(0, 1))  # clamped, not extrapolated
```

### Triangular: `quantile_triangular()`

`quantile_triangular(fit, q)` takes `fit = list(min=, mode=, max=)` and evaluates the standard
closed-form triangular inverse-CDF, split at the mode fraction `fc = (mode - min) / (max - min)`.
Here the l/r/h triplet plays min/mode/max directly:

```{r}
triangular_fit <- list(min = l, mode = r, max = h)
quantile_triangular(triangular_fit, q = c(0.05, 0.5, 0.95))
```

### Metalog: `fit_metalog_linear()` / `quantile_metalog_linear()`

With exactly 3 interior percentiles and 3 metalog terms, `fit_metalog_linear()` solves an
exactly-determined linear system for the metalog coefficients (`solve(Y, z)`) rather than
optimizing. `boundedness = "u"` (unbounded) is used here since no physical `bounds` are supplied.

```{r}
metalog_fit <- fit_metalog_linear(interior_values = c(l, r, h), interior_probs = c(0.05, 0.5, 0.95),
                                   boundedness = "u")
quantile_metalog_linear(metalog_fit, q = c(0.05, 0.5, 0.95))  # exact at the knots, like the others
```

### Comparing all five fitted PDFs on the same plot

Every family above reproduces `l`/`r`/`h` exactly at `q = 0.05/0.5/0.95` (that's what "fitting to
the triplet" means for each of them) - what differs is the *shape* connecting those three points.
Evaluating each quantile function on a fine probability grid and differentiating numerically
approximates each family's density:

```{r, fig.height = 4.5}
q_grid <- seq(0.01, 0.99, by = 0.01)

families <- list(
  Normal     = quantile_normal(normal_fit, q_grid),
  Beta       = bounds[1] + quantile_beta(beta_fit, q_grid) * diff(bounds),
  LinearCDF  = quantile_linear_cdf(c(0.05, 0.5, 0.95), c(l, r, h), q_grid),
  Triangular = quantile_triangular(triangular_fit, q_grid),
  Metalog    = quantile_metalog_linear(metalog_fit, q_grid)
)

cdf_df <- do.call(rbind, lapply(names(families), function(fam) {
  data.frame(family = fam, q = q_grid, value = families[[fam]])
}))

ggplot(cdf_df, aes(x = value, y = q, color = family)) +
  geom_line(linewidth = 1) +
  scale_color_viridis_d(name = "Family") +
  theme_minimal() +
  labs(
    title = "Fitted CDFs of the same l/r/h triplet, five families",
    subtitle = "l = 15, r = 25, h = 40 at p = 0.05 / 0.50 / 0.95",
    x = "Value (clay %)", y = "Cumulative probability"
  )
```

The families agree exactly at the three fitted knots (0.05/0.5/0.95) but diverge everywhere else -
Triangular is angular by construction, Normal/Beta/Metalog are smooth but differently shaped in the
tails, and LinearCDF is piecewise-straight between knots with hard clamping beyond them (visible as
the flat segments at the very top/bottom of its curve).

## Step 2: Automatic family selection - `resolve_property_family()`

`resolve_property_family(l, r, h, bounds = NULL, skew_threshold = 0.15)` picks a family from a skew
proxy `((h - r) - (r - l)) / (h - l)` - how lopsided the high-side spread is relative to the
low-side spread around the representative value. If `bounds` is supplied it always resolves to
`"beta"`; otherwise `"normal"` if the skew proxy is within `skew_threshold` of zero, else
`"lognormal"`. It deliberately never resolves to `"metalog"` - that stays an explicit, opted-in
choice.

```{r}
symmetric  <- list(l = 15, r = 25, h = 35)   # (35-25) - (25-15) = 0            -> skew_proxy 0
mild_skew  <- list(l = 15, r = 25, h = 40)   # (40-25) - (25-15) = 5            -> skew_proxy 0.25
strong_skew <- list(l = 20, r = 24, h = 60)  # (60-24) - (24-20) = 32           -> skew_proxy 0.9

triplets <- list(Symmetric = symmetric, `Mild skew` = mild_skew, `Strong skew` = strong_skew)

resolution_df <- do.call(rbind, lapply(names(triplets), function(nm) {
  t <- triplets[[nm]]
  res <- resolve_property_family(t$l, t$r, t$h)
  data.frame(triplet = nm, l = t$l, r = t$r, h = t$h,
             skew_proxy = round(res$skew_proxy, 3), family = res$family)
}))
resolution_df
```

The symmetric triplet resolves to `"normal"` (skew proxy exactly 0, well inside the default 0.15
threshold); both skewed triplets resolve to `"lognormal"`. Sliding `skew_threshold` up past a
triplet's own skew proxy flips the decision from `"lognormal"` back to `"normal"`:

```{r}
thresholds <- c(0.05, 0.15, 0.30, 0.50)
threshold_df <- do.call(rbind, lapply(thresholds, function(th) {
  res <- resolve_property_family(mild_skew$l, mild_skew$r, mild_skew$h, skew_threshold = th)
  data.frame(skew_threshold = th, skew_proxy = round(res$skew_proxy, 3), family = res$family)
}))
threshold_df
```

The mild-skew triplet's skew proxy is 0.25; once `skew_threshold` rises above that (0.30, 0.50), the
same triplet resolves to `"normal"` instead of `"lognormal"` - the same input data, a different
family, purely from where the threshold is drawn. Supplying `bounds` overrides all of this and
always resolves to `"beta"`:

```{r}
resolve_property_family(mild_skew$l, mild_skew$r, mild_skew$h, bounds = c(0, 100))$family
```

## Step 3: `fit_percentile_triplet()` - the master dispatcher

`fit_percentile_triplet(l, r, h, family, lh_probs = c(0.05, 0.95), bounds = NULL, boundedness = ...)`
is the single entry point the rest of soilSIM calls - it wraps every fitter shown in Step 1 behind
one `family` argument (`"triangular"`, `"uniform"`, `"normal"`, `"lognormal"`, `"beta"`, `"metalog"`,
`"linear_cdf"`, or `"auto"`), plus two degenerate-input guards: non-finite `l`/`r`/`h` returns
`valid = FALSE`, and a zero-width interval (`h == l`) always collapses to a triangular fit
`min = mode = max = r` regardless of the requested family, since every parametric family divides by
zero on a zero-width interval.

```{r}
for (fam in c("normal", "beta", "lognormal", "triangular", "metalog", "auto")) {
  result <- fit_percentile_triplet(l, r, h, family = fam, bounds = if (fam == "beta") bounds else NULL)
  cat(sprintf("%-10s -> resolved family = %-10s valid = %s\n", fam, result$family, result$valid))
}
```

`family = "auto"` with no `bounds` resolves through `resolve_property_family()` exactly as in Step
2 (this triplet's skew proxy of 0.25 exceeds the default 0.15 threshold, so it resolves to
`"lognormal"`). Once fitted, `quantile_from_fit(u, family, fit)` is the matching master *quantile*
dispatcher - the same round-trip Step 1 built by hand for each family individually:

```{r}
auto_fit <- fit_percentile_triplet(l, r, h, family = "auto")
quantile_from_fit(u = c(0.05, 0.5, 0.95), family = auto_fit$family, fit = auto_fit$fit)
```

The zero-width-interval guard, demonstrated directly:

```{r}
fit_percentile_triplet(l = 25, r = 25, h = 25, family = "beta", bounds = bounds)
```

## Step 4: `fit_beta_mle_newton()`'s `n_iter` parameter - watching convergence

`fit_beta_mle_newton(values, bounds, n_iter = 15, eps = 1e-6)` runs a *fixed* number of
Newton-Raphson steps - there is no internal convergence check, so `n_iter` directly controls how
close the fit gets to the true MLE. Refitting the same values at increasing `n_iter` shows the
shape parameters (and the resulting quantiles) settle down after only a few steps:

```{r}
n_iters <- c(1, 2, 3, 5, 10, 15)
convergence_df <- do.call(rbind, lapply(n_iters, function(ni) {
  fit_i <- fit_beta_mle_newton(values = c(l, r, h), bounds = bounds, n_iter = ni)
  q_i <- bounds[1] + quantile_beta(fit_i, q = c(0.05, 0.5, 0.95)) * diff(bounds)
  data.frame(n_iter = ni, shape1 = fit_i$shape1, shape2 = fit_i$shape2,
             q05 = q_i[1], q50 = q_i[2], q95 = q_i[3])
}))
convergence_df
```

```{r}
ggplot(convergence_df, aes(x = n_iter, y = shape1)) +
  geom_line(color = viridisLite::viridis(1, begin = 0.2), linewidth = 1) +
  geom_point(aes(color = n_iter), size = 3) +
  scale_color_viridis_c(guide = "none") +
  theme_minimal() +
  labs(
    title = "fit_beta_mle_newton(): shape1 converges within a few Newton-Raphson steps",
    x = "n_iter", y = "Fitted shape1"
  )
```

`shape1` (and `shape2`, not plotted) moves the most between `n_iter = 1` and `n_iter = 3`, then
barely changes from `n_iter = 5` onward - `n_iter = 15` (soilSIM's default everywhere else) is well
past the point of diminishing returns for this triplet, which is exactly why a fixed iteration count
without a convergence check is a safe design choice here.

## Step 5: ILR compositional transforms

Soil texture (clay/sand/silt) is a 3-part composition that must sum to a fixed total (e.g. 100%) and
stay non-negative - ordinary independent simulation of each fraction can't guarantee that.
`ilr_forward()`/`ilr_inverse()` are soilSIM's isometric log-ratio transform: they move a composition
to/from a 2-dimensional unconstrained coordinate system where ordinary Normal-distribution machinery
applies safely, with the constraint automatically satisfied on the way back.

```{r}
# A synthetic clay/sand/silt composition
example_composition <- c(clay = 24, sand = 40, silt = 36)
sum(example_composition)

z <- ilr_forward(clay = example_composition["clay"], sand = example_composition["sand"],
                  silt = example_composition["silt"])
z  # 2 unconstrained coordinates, z1/z2
```

```{r}
back <- ilr_inverse(z1 = z[, "z1"], z2 = z[, "z2"], total = 100)
back  # round-trips back to the original composition
sum(back)
```

Perturbing the ILR coordinates and inverting always yields a valid composition - non-negative and
summing exactly to `total` - which is the point of doing simulation in ILR space rather than
directly on clay/sand/silt:

```{r, fig.height = 4.5}
set.seed(42)
z1_perturbed <- z[, "z1"] + seq(-1.5, 1.5, length.out = 9)
z2_perturbed <- rep(z[, "z2"], 9)
perturbed_compositions <- as.data.frame(ilr_inverse(z1_perturbed, z2_perturbed, total = 100))
perturbed_compositions$step <- seq_len(9)

stack_df <- do.call(rbind, lapply(c("clay", "sand", "silt"), function(part) {
  data.frame(step = perturbed_compositions$step, fraction = part,
             pct = perturbed_compositions[[part]])
}))

ggplot(stack_df, aes(x = step, y = pct, fill = fraction)) +
  geom_col() +
  scale_fill_viridis_d(name = NULL) +
  theme_minimal() +
  labs(
    title = "ilr_inverse(): every reconstructed composition stays valid",
    subtitle = "z1 swept across a range at fixed z2 - always sums to 100, always non-negative",
    x = "Step (increasing z1)", y = "Percent"
  )
```

Every bar in the stack sums to exactly 100 - `ilr_inverse()`'s softmax-like normalization guarantees
this by construction, for any finite `z1`/`z2`.

## Step 6: Correlation-matrix repair

Multi-property Monte Carlo simulation needs a positive-definite (PD) correlation matrix for its
Cholesky-copula step. Empirically estimated or hand-assembled correlation matrices are not always
PD - `ensure_positive_definite_matrix()` repairs one via eigenvalue flooring, and
`validate_correlation_matrix()` checks whether a candidate matrix is already usable.

A deliberately broken, non-positive-definite 3x3 correlation matrix (valid-looking diagonal and
range, but internally inconsistent pairwise correlations):

```{r}
broken_matrix <- matrix(
  c( 1.0,  0.9, -0.9,
     0.9,  1.0,  0.9,
    -0.9,  0.9,  1.0),
  nrow = 3, dimnames = list(c("clay", "sand", "silt"), c("clay", "sand", "silt"))
)
broken_matrix
eigen(broken_matrix, only.values = TRUE)$values  # a negative eigenvalue confirms it's not PD
```

```{r}
validate_correlation_matrix(broken_matrix, properties = c("clay", "sand", "silt"))
```

`ensure_positive_definite_matrix(matrix, min_eigenvalue = 1e-6)` eigendecomposes the matrix, floors
any eigenvalue below `min_eigenvalue`, reconstructs, and rescales back to a unit-diagonal
correlation matrix (also restoring the original dimnames, which `eigen()$vectors` does not carry):

```{r}
repaired_matrix <- ensure_positive_definite_matrix(broken_matrix)
repaired_matrix
eigen(repaired_matrix, only.values = TRUE)$values  # all positive now
validate_correlation_matrix(repaired_matrix, properties = c("clay", "sand", "silt"))
```

Raising `min_eigenvalue` pushes the repaired matrix further from the (broken) original toward a
better-conditioned but less "confident" correlation structure - visible as the off-diagonal
correlations shrinking toward zero as the floor rises:

```{r, fig.height = 4.5}
floors <- c(1e-6, 0.01, 0.1, 0.3)
floor_df <- do.call(rbind, lapply(floors, function(mv) {
  r <- ensure_positive_definite_matrix(broken_matrix, min_eigenvalue = mv)
  data.frame(min_eigenvalue = mv, clay_sand = r["clay", "sand"], clay_silt = r["clay", "silt"],
             sand_silt = r["sand", "silt"])
}))
floor_df_long <- do.call(rbind, lapply(c("clay_sand", "clay_silt", "sand_silt"), function(pair) {
  data.frame(min_eigenvalue = floor_df$min_eigenvalue, pair = pair, correlation = floor_df[[pair]])
}))

ggplot(floor_df_long, aes(x = factor(min_eigenvalue), y = correlation, color = pair, group = pair)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  scale_color_viridis_d(name = "Property pair") +
  theme_minimal() +
  labs(
    title = "ensure_positive_definite_matrix(): higher min_eigenvalue shrinks correlations",
    x = "min_eigenvalue", y = "Repaired correlation"
  )
```

`validate_correlation_matrix(corr_matrix, properties)` is the cheaper, non-repairing counterpart -
it only checks shape (square, matching `length(properties)`), unit diagonal, and positive
definiteness, returning the first failing check's message rather than fixing anything. Together
these two functions are the shared final guard both `estimate_correlation_matrix_robust()` and
`build_kssl_fallback_matrix()` route their correlation matrices through before the Cholesky-copula
simulation step consumes them.
