---
title: "GP Depth Modeling & Multivariate Adjustment: Function by Function"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GP Depth Modeling & Multivariate Adjustment: Function by Function}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6, fig.height = 4)
```

## Overview

`getting-started-monte-carlo.Rmd` walks the full tabular pipeline end to end, and its final step
only touches Gaussian-process (GP) depth modeling briefly - fitting one set of stratified GP models
and plotting a single depth trend. This vignette is a companion, finer-grained tour of the same
functional group: it walks through the *individual* functions in `R/gp-modeling.R` and
`R/multivariate-adjustment.R` one at a time, explains what each parameter controls (not just how to
call the function), and shows parameter effects visually. See
`soilSIM/docs/05_gp_modeling_multivariate_adjustment.md` for the full function-level architecture
reference this vignette is drawn from.

Functions covered, in order:

1. `prepare_nrcs_training_data()` - shape raw NRCS/SSURGO data into GP training data
2. `select_optimal_grouping()` - choose a soil-grouping strategy for stratified GP fitting
3. `build_stratified_gp_models()` and `fit_individual_gp_model()` - fit the depth-trend GPs
4. `predict_gp_depth_trends()` - predict a fitted model at new depths, with an uncertainty band
5. `validate_gp_models()` - diagnostic pass over a fitted model set
6. `adjust_multivariate_depthwise_GP()` - why correlation-preserving adjustment matters
7. `preserve_correlation_structure()` / `apply_gp_depth_trends()` - the long-format production path
8. `apply_nrcs_trend_adjustments()` vs. `apply_local_gp_adjustments()` - regional vs. local trends
9. `preserve_correlation_structure_joint()` - the joint depth x property copula method (now the
   package default), A/B tested against the original `gp_quantile_retrofit` algorithm from Step 7
   above on this vignette's own real fitted GP models

This vignette reuses the same real, cached Amador County, CA SSURGO data
(`inst/extdata/ssurgo_amador.rds`) that `getting-started-monte-carlo.Rmd` uses, following the same
preprocessing steps, so results here are directly comparable to that vignette.

```{r setup}
library(soilSIM)
library(ggplot2)
```

## Step 0: Get to the same starting point as the Getting Started vignette

`prepare_nrcs_training_data()` (below) takes the same *processed but not-yet-infilled* SSURGO
horizon data that `getting-started-monte-carlo.Rmd` calls `horizon_data` - the output of
`process_ssurgo_data()`, before any missing-value infilling. `download_and_prepare_ssurgo()` is the
real, live call that produced the cached data (shown for reference only - not run here):

```{r, eval = FALSE}
# The real call the cached data came from:
ssurgo_amador <- download_and_prepare_ssurgo(
  aoi_wkt = "POLYGON((-120.5 38.5, -120.4 38.5, -120.4 38.6, -120.5 38.6, -120.5 38.5))",
  properties = c("sandtotal", "claytotal", "silttotal", "dbovendry", "ph1to1h2o",
                 "cec7", "om", "wthirdbar", "wfifteenbar"),
  max_depth = 150
)
```

```{r}
ssurgo_amador <- readRDS(system.file("extdata", "ssurgo_amador.rds", package = "soilSIM"))
raw_data <- ssurgo_amador$ssurgo_data

processed <- process_ssurgo_data(raw_data, max_depth = 150, verbose = FALSE)
horizon_data <- processed$processed_data
dim(horizon_data)
```

## Step 1: `prepare_nrcs_training_data()` - shaping raw data for GP fitting

**Purpose**: turn raw combined NRCS component/horizon data into a cleaned, grouped training set
ready for `build_stratified_gp_models()`.

**Parameters that matter**:

- `grouping_strategy` - `"auto"` (the default here) hands the decision to
  `select_optimal_grouping()` (Step 2); or force one of `"soil_series"`, `"taxonomic_class"`,
  `"particle_size"`, `"soil_grtgroup"`, `"soil_suborder"`, `"soil_order"`, `"none"` directly.
- `min_profiles_per_group` / `min_observations_per_group` - how many distinct components / total
  horizon rows a group needs before it's considered adequate for GP fitting (used both by the
  auto-selector and to drop inadequate groups from the final output).
- `target_min_groups` - how many adequate groups the auto-selector wants to find before it accepts
  a strategy (higher values push it toward coarser, more-populated groupings).
- `max_depth` - horizons deeper than this (cm) are dropped before fitting.

Internally it also validates data quality, standardizes property names, flags and removes
unsuitable horizons (bedrock, cemented pans, etc.), and coalesces synonym columns
(`claytotal_r`/`clay_r`/`clay` -> `clay_pct`, and similarly for the other seven standard
properties) - see the doc for the full algorithm.

```{r}
gp_train <- prepare_nrcs_training_data(horizon_data, max_depth = 150)
dim(gp_train)
table(gp_train$soil_group)
```

The output has the eight standardized property columns (`clay_pct`, `sand_pct`, `silt_pct`, `pH`,
`organic_matter`, `bulk_density`, `cec`, `awc`), plus `soil_group` and `depth_midpoint`, restricted
to groups meeting `min_observations_per_group`.

## Step 2: `select_optimal_grouping()` - choosing a grouping strategy

`prepare_nrcs_training_data()` calls this internally when `grouping_strategy = "auto"`, but it's a
useful function to call directly to see *why* a particular strategy was chosen. It walks a fixed
hierarchy - `soil_series -> taxonomic_class -> particle_size -> soil_grtgroup -> soil_suborder ->
soil_order -> none` - skipping any strategy whose required column is missing, and returns the first
strategy that produces at least `target_groups` "adequate" groups (`>= min_profiles` distinct
components, `>= min_obs` rows, and `>= 20` cm of depth range). If no strategy meets the target, it
falls back to whichever produced the most adequate groups.

`target_groups` is the parameter that most directly controls how fine- or coarse-grained the
resulting grouping is - raising it forces the selector further down the hierarchy (coarser
groupings, since coarser strategies pool more components together):

```{r}
select_optimal_grouping(horizon_data, min_profiles = 3, min_obs = 15, target_groups = 3)
select_optimal_grouping(horizon_data, min_profiles = 3, min_obs = 15, target_groups = 15)
```

For this small AOI, asking for far more adequate groups than the data can really support
(`target_groups = 15`) either falls back to a coarser strategy or falls back to the
best-available option with a warning logged - a direct illustration of `target_groups` trading off
grouping resolution against data adequacy per group.

## Step 3: Fitting depth-trend GP models

### `build_stratified_gp_models()` - the master fitting function

**Concept**: for each requested property, and separately within each soil group, this fits a 1-D
Gaussian Process regression of the property against (scaled) depth - i.e., a smooth, flexible
depth-trend curve with uncertainty, rather than a single depth-independent mean. Groups that are too
small get folded into progressively coarser fallback groups (particle size -> great group ->
suborder -> order -> a general pool) via `apply_hierarchical_grouping()` before fitting, so every
group that reaches `fit_individual_gp_model()` has enough training points.

**Parameters that matter**:

- `properties` - which standardized property columns to model.
- `min_profiles_per_group` / `min_observations_per_group` - adequacy thresholds used by the
  fallback grouping (same meaning as in Step 1).
- `optimize_hyperparameters` - `TRUE` runs `optimize_gp_hyperparameters()`, a 5-fold
  cross-validated search over three `GPfit` correlation families (exponential power, Matern
  nu=3/2, Matern nu=5/2), picking the lowest mean held-out RMSE; `FALSE` fits a single default GP
  directly. `FALSE` is dramatically faster and is what this vignette uses (matching
  `getting-started-monte-carlo.Rmd`'s Step 5), at the cost of not searching over correlation
  families.

```{r}
gp_models <- build_stratified_gp_models(
  gp_train,
  properties = c("clay_pct", "sand_pct", "pH", "organic_matter"),
  min_profiles_per_group = 3,
  min_observations_per_group = 15,
  optimize_hyperparameters = FALSE
)
gp_models$model_summary$total_models
names(gp_models$clay_pct$models)
```

`r gp_models$model_summary$total_models` GP models were fit in total, across
`r length(gp_models$clay_pct$models)` clay-content groups (and similarly for the other three
properties) from this AOI's real data.

### `fit_individual_gp_model()` - the per-group fit, called directly

This is the function `build_stratified_gp_models()` calls once per (property, group) combination.
Calling it directly on a single group's data shows what it actually does: aggregate to one mean
value per depth (across profiles, then across horizons at that depth), min-max scale depths to
`[0, 1]` (so the GP's length-scale hyperparameter is on a comparable footing across groups with
different depth ranges), and fit `GPfit::GP_fit()`. It returns `NULL` rather than erroring when
there's too little data to fit (fewer than 3 aggregated depth points, no variance in values, or no
depth range) - `build_stratified_gp_models()` relies on this to skip ungroupable data gracefully.

```{r}
# Pick a soil_group that already has a successfully-fit clay_pct AND
# organic_matter model in gp_models (used again in later steps), with the
# most training points among those - guarantees the direct, un-fallback-ed
# fit_individual_gp_model() call below actually succeeds (fitting can return
# NULL for groups whose raw soil_group data - as opposed to
# build_stratified_gp_models()'s internal hierarchical-fallback final_group -
# turns out to have no depth variation for this property).
candidate_groups <- intersect(names(gp_models$clay_pct$models), names(gp_models$organic_matter$models))
candidate_groups <- candidate_groups[order(
  sapply(gp_models$clay_pct$models[candidate_groups], function(m) m$n_training_points),
  decreasing = TRUE
)]

single_fit <- NULL
for (candidate in candidate_groups) {
  group_data <- gp_train[gp_train$soil_group == candidate, ]
  single_fit <- fit_individual_gp_model(group_data, "clay_pct", optimize_hyperparameters = FALSE)
  if (!is.null(single_fit)) {
    demo_group <- candidate
    break
  }
}
demo_group
single_fit$model$depth_scaling
single_fit$model$n_training_points
single_fit$diagnostics
```

`depth_scaling` (the training depth range, min-max, rescaled to `[0, 1]`) is stored on the fitted
model precisely so that `predict_gp_depth_trends()` (next step) can rescale *new* depths the same
way before calling `GPfit::predict.GP()`.

## Step 4: `predict_gp_depth_trends()` - depth-trend curves with a prediction interval

**Purpose**: predict a fitted GP model's property values at arbitrary new depths. It rescales
`new_depths` into the `[0, 1]` range the model was trained on (using the stored `depth_scaling`),
clamps to `[0, 1]` (so predictions outside the training depth range are extrapolated at the
boundary rather than on a nonsensical negative/> 1 scale), and calls `GPfit::predict.GP()`.

`predict_gp_depth_trends()` itself only returns the fitted mean (`Y_hat`) - not the prediction
uncertainty. To draw an uncertainty band, pull the prediction variance directly from
`GPfit::predict.GP()` using the same rescaling logic, exactly as `getting-started-monte-carlo.Rmd`
does in its final step. Here for a different property/group than that vignette used - `organic_matter`
in this same `demo_group`:

```{r}
om_gp_model <- gp_models$organic_matter$models[[demo_group]]

depth_seq <- seq(0, 150, by = 5)
om_mean <- predict_gp_depth_trends(om_gp_model, depth_seq)

scaling <- om_gp_model$depth_scaling
scaled_depths <- pmax(0, pmin(1, (depth_seq - scaling$min) / scaling$range))
om_pred <- GPfit::predict.GP(om_gp_model$gp_model, xnew = as.matrix(scaled_depths))
om_se <- sqrt(pmax(om_pred$MSE, 0))

om_depth_trend <- data.frame(
  depth = depth_seq,
  mean = om_mean,
  lower = om_mean - 1.96 * om_se,
  upper = om_mean + 1.96 * om_se
)

ggplot(om_depth_trend, aes(x = depth, y = mean)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = viridisLite::viridis(1, begin = 0.5), alpha = 0.25) +
  geom_line(color = viridisLite::viridis(1, begin = 0.15), linewidth = 1) +
  theme_minimal() +
  labs(
    title = paste0("GP depth trend: organic matter, \"", demo_group, "\" soil group"),
    subtitle = "Shaded band is a 95% prediction interval",
    x = "Depth (cm)",
    y = "Organic matter (%)"
  )
```

The band narrows near depths with more training observations and widens where the GP is
extrapolating - exactly the behavior a GP is meant to provide over a plain regression trend line.

## Step 5: `validate_gp_models()` - diagnosing a fitted model set

**Purpose**: a diagnostic pass over an entire fitted GP model set (the output of
`build_stratified_gp_models()`), checking three things per (property, group) model: whether
predictions come back at all (`model_valid`), whether the depth trend is reasonably monotonic
(`>= 70%` of consecutive depth-to-depth differences share the same sign, via
`assess_trend_monotonicity()`), and whether predictions fall inside a property-specific realistic
range (e.g. clay/sand/silt in `[0, 100]`, pH in `[2.5, 11]`, via `assess_realistic_values()`).

```{r}
validation <- validate_gp_models(gp_models, nrcs_data = gp_train, validation_depths = seq(0, 150, by = 10))
validation$overall_assessment
validation$recommendations
```

`overall_assessment$validation_passed` is `TRUE` when `>= 80%` of all fitted models pass; anything
lower is worth investigating via `model_validation` before trusting the model set downstream.

## Step 6: Why correlation-preserving adjustment matters

Once GP depth-trend models exist, the next step in the full pipeline (see
`getting-started-monte-carlo.Rmd`'s closing note and
`soilSIM/docs/05_gp_modeling_multivariate_adjustment.md`) is to nudge Monte Carlo simulation
realizations toward those depth trends. The naive way to do this - adjust each property's
realizations independently to match its own GP trend - breaks the *cross-property* correlation that
the Monte Carlo engine worked to preserve in the first place (e.g. clay and sand content are
naturally anti-correlated within a horizon; adjusting each to its own trend independently reshuffles
which simulated clay value is paired with which simulated sand value). `adjust_multivariate_depthwise_GP()`
avoids this by computing a single *reference quantile* per simulation (from the primary property's
surface-depth distribution) and using that same quantile - not each property's own independent
quantile - to drive every property's adjustment, so realization *j*'s properties all move together.

To demonstrate concretely with real fitted models, build a small synthetic set of "before
adjustment" Monte Carlo-style realizations for clay and sand content at a few depths, correlated
within each simulation (mimicking what `generate_monte_carlo_realizations()` produces before any
depth-trend adjustment is applied) but with no depth trend yet:

```{r}
set.seed(2024)
n_sims <- 300
demo_depths <- c(0, 20, 40, 60, 80)

common_groups <- intersect(names(gp_models$clay_pct$models), names(gp_models$sand_pct$models))
xv_group <- common_groups[1]
clay_gp_model <- gp_models$clay_pct$models[[xv_group]]
sand_gp_model <- gp_models$sand_pct$models[[xv_group]]

# One shared per-simulation draw, reused at every depth (no trend yet) - clay and
# sand are drawn jointly and negatively correlated, as real compositional
# texture properties are.
z_clay <- rnorm(n_sims)
target_rho <- -0.65
z_sand <- target_rho * z_clay + sqrt(1 - target_rho^2) * rnorm(n_sims)

clay_base <- mean(clay_gp_model$training_data$mean_value)
sand_base <- mean(sand_gp_model$training_data$mean_value)

clay_mat <- matrix(clay_base + 6 * z_clay, nrow = length(demo_depths), ncol = n_sims, byrow = TRUE)
sand_mat <- matrix(sand_base + 8 * z_sand, nrow = length(demo_depths), ncol = n_sims, byrow = TRUE)

simulated_list <- list(clay_pct = clay_mat, sand_pct = sand_mat)
cor(clay_mat[1, ], sand_mat[1, ])  # starting correlation, ~ target_rho
```

Now adjust toward the two real fitted GP depth trends two different ways: **jointly** (the
correlation-preserving way `adjust_multivariate_depthwise_GP()` is designed for) and **naively**,
by calling the same function separately on each property in isolation - which forces each property
to use *its own* surface distribution as the reference quantile ordering instead of sharing one:

```{r}
joint_adjusted <- adjust_multivariate_depthwise_GP(
  simulated_list = simulated_list,
  gp_models = list(clay_pct = clay_gp_model, sand_pct = sand_gp_model),
  depths = demo_depths,
  primary_property = "clay_pct"
)

naive_clay <- adjust_multivariate_depthwise_GP(
  simulated_list = simulated_list["clay_pct"],
  gp_models = list(clay_pct = clay_gp_model),
  depths = demo_depths
)

naive_sand <- adjust_multivariate_depthwise_GP(
  simulated_list = simulated_list["sand_pct"],
  gp_models = list(sand_pct = sand_gp_model),
  depths = demo_depths
)

deep_idx <- length(demo_depths)

cor_comparison <- data.frame(
  stage = factor(
    c("Original\n(unadjusted)", "Naive\n(each property adjusted alone)", "Correlation-preserving\n(adjust_multivariate_depthwise_GP)"),
    levels = c("Original\n(unadjusted)", "Naive\n(each property adjusted alone)", "Correlation-preserving\n(adjust_multivariate_depthwise_GP)")
  ),
  correlation = c(
    cor(clay_mat[deep_idx, ], sand_mat[deep_idx, ]),
    cor(naive_clay$clay_pct[deep_idx, ], naive_sand$sand_pct[deep_idx, ]),
    cor(joint_adjusted$clay_pct[deep_idx, ], joint_adjusted$sand_pct[deep_idx, ])
  )
)
cor_comparison
```

```{r}
ggplot(cor_comparison, aes(x = stage, y = correlation, fill = correlation)) +
  geom_col(width = 0.6) +
  geom_hline(yintercept = target_rho, linetype = "dashed", color = "grey40") +
  scale_fill_viridis_c(limits = c(-1, 1)) +
  theme_minimal() +
  labs(
    title = "Clay-sand correlation at the deepest demo depth",
    subtitle = "Dashed line is the original correlation before any depth-trend adjustment",
    x = NULL, y = "Correlation (clay_pct, sand_pct)"
  ) +
  theme(legend.position = "none")
```

Adjusting each property in isolation lets its quantile ordering drift away from the other
property's - clay's deep-depth quantile ranking no longer lines up with sand's, so their
correlation degrades from the original. The joint, correlation-preserving call keeps both
properties keyed to the *same* per-simulation reference quantile at every depth, so the original
correlation survives the depth-trend adjustment nearly intact.

## Step 7: `preserve_correlation_structure()` / `apply_gp_depth_trends()` - the long-format path

`adjust_multivariate_depthwise_GP()` above (from `gp-modeling.R`) operates directly on
`[depth x simulation]` matrices and is the original, array-based implementation of this algorithm.
The pipeline `integrate_monte_carlo_with_gp()` actually calls works on long-format Monte Carlo
output instead, via `multivariate-adjustment.R`'s `apply_gp_depth_trends()` (reshapes long-format
rows into matrices, dispatches to `preserve_correlation_structure()` when
`preserve_correlations = TRUE`, at least two properties are present, and
`config$monte_carlo$vertical_correlation_method = "gp_quantile_retrofit"` is explicitly set - see
Step 9 below for the package's actual current default, `"joint_copula"` - then reshapes back) and
`preserve_correlation_structure()` itself (the identical reference-quantile algorithm - same
`calculate_safe_gp_ratio()` clamped-ratio step, same `apply_quantile_adjustment()` reference-quantile
nudge, same `correct_distribution_shape()` marginal-distribution remap - just operating on
`property_matrices`/`gp_predictions` named lists rather than being handed `gp_models` objects
directly). `apply_gp_depth_trends()` is used indirectly below, via `apply_nrcs_trend_adjustments()`
and `apply_local_gp_adjustments()`.

## Step 8: NRCS-generalized vs. local models

Two different sources of "what depth trend should this cokey be nudged toward" exist in the
package, and they're used via two different functions:

- **`apply_nrcs_trend_adjustments()`** - looks up the cokey's matched *regional* GP model group
  (from `match_soils_to_gp_models()`, e.g. "this cokey's soil series") in the stratified GP model
  set fit in Step 3, and uses *that* group's depth trend - fit from potentially many profiles
  across the AOI. It "borrows strength" from the broader dataset, at the cost of being generic to
  the group rather than specific to this one cokey.
- **`apply_local_gp_adjustments()`** - fits a brand-new, tiny GP (via `fit_local_gp_models()`)
  using *only* the current cokey's own aggregated-by-depth data (`min_depths = 3` unique depths
  required). It captures whatever depth pattern this specific cokey's own horizons show, but with
  very few training points it is much more sensitive to noise in that one profile.

`integrate_monte_carlo_with_gp(integration_method = "hybrid")` (the default) actually applies both,
in sequence, per cokey. Demonstrate the two independently on the synthetic `demo_cokey_data`
built from Step 6's matrices (which, by construction, has *no* real depth trend - just noise around
a constant mean at each depth - so any trend either function introduces is coming entirely from the
adjustment itself):

```{r}
demo_cokey_data <- expand.grid(hzdept_r = demo_depths, simulation_number = seq_len(n_sims))
demo_cokey_data$claytotal <- as.vector(clay_mat)  # column-major flatten matches expand.grid's order
demo_cokey_data$sandtotal <- as.vector(sand_mat)
demo_cokey_data$cokey <- "demo_cokey"

nrcs_adjusted <- apply_nrcs_trend_adjustments(
  cokey_data = demo_cokey_data,
  gp_models = gp_models,
  model_group = xv_group,
  properties = c("claytotal", "sandtotal")
)

local_adjusted <- apply_local_gp_adjustments(
  cokey_data = demo_cokey_data,
  properties = c("claytotal", "sandtotal")
)
```

Compare the resulting mean clay-content-by-depth curves against the real regional GP trend that
`apply_nrcs_trend_adjustments()` pulled from (`clay_gp_model`, fit from many profiles in Step 3) and
against the flat, untrended original:

```{r}
regional_trend <- data.frame(
  depth = demo_depths,
  mean_clay = predict_gp_depth_trends(clay_gp_model, demo_depths),
  source = "Regional GP trend (this cokey's matched group)"
)

trend_comparison <- rbind(
  regional_trend,
  data.frame(
    depth = demo_depths,
    mean_clay = tapply(demo_cokey_data$claytotal, demo_cokey_data$hzdept_r, mean),
    source = "Original (untrended synthetic data)"
  ),
  data.frame(
    depth = demo_depths,
    mean_clay = tapply(nrcs_adjusted$claytotal, nrcs_adjusted$hzdept_r, mean),
    source = "apply_nrcs_trend_adjustments() (regional model)"
  ),
  data.frame(
    depth = demo_depths,
    mean_clay = tapply(local_adjusted$claytotal, local_adjusted$hzdept_r, mean),
    source = "apply_local_gp_adjustments() (this cokey only)"
  )
)

ggplot(trend_comparison, aes(x = depth, y = mean_clay, color = source)) +
  geom_line(linewidth = 1) +
  geom_point() +
  scale_color_viridis_d(end = 0.9) +
  theme_minimal() +
  labs(
    title = "Regional vs. local GP depth-trend adjustment",
    subtitle = "Synthetic demo cokey with no inherent depth trend",
    x = "Depth (cm)", y = "Mean simulated clay content (%)", color = NULL
  ) +
  theme(legend.position = "bottom", legend.direction = "vertical")
```

`apply_nrcs_trend_adjustments()` pulls the mean curve toward the regional GP trend fit from many
profiles in Step 3. `apply_local_gp_adjustments()`, fitting a GP from only this cokey's own 5
(noisy, untrended) depth means, tends to track that noise rather than impose a strong trend - useful
when a cokey's own data really does deviate from its group's regional pattern, but a weaker signal
to lean on than the regional model when a cokey has little data of its own.

## Step 9: why the production default changed - the joint depth x property copula

Steps 6-7 above walked through `preserve_correlation_structure()` - the ORIGINAL algorithm this
package used, still fully supported today as an explicit opt-out
(`config$monte_carlo$vertical_correlation_method = "gp_quantile_retrofit"`). It keeps
cross-*property* correlation intact while nudging realizations toward a GP-fitted depth trend, via
a shared reference-quantile trick. That same mechanism has a side effect that motivated replacing
it as the default: because it anchors EVERY depth's adjustment to one FIXED per-realization
percentile (established once, at the surface, and reused unchanged at every deeper depth), it
doesn't just preserve realization *ranking* - it forces realization *j* to occupy the same
percentile at every depth, for the life of the profile. That is a much stronger, more rigid
constraint than "vertical correlation": it's independent of how quickly or slowly the underlying
property actually varies with depth (the GP model's own fitted smoothness plays no role here at
all), and - as the comparison below shows on this vignette's own real fitted models - the result
is spuriously near-perfect (`> 0.99`) correlation across the *entire* depth profile, not a
realistic decay.

`VERTICAL_CORRELATION_IMPROVEMENT_PLAN.md` (in the package root) documents the replacement:
`preserve_correlation_structure_joint()`, wired into `apply_gp_depth_trends()` behind
`config$monte_carlo$vertical_correlation_method` (now defaulting to `"joint_copula"` as of that
plan's Phase 13 - `"gp_quantile_retrofit"` remains available as an explicit opt-out, not removed).
Instead of a sequential mean-nudge, it draws depth correlation and property correlation
**simultaneously** from a single Kronecker-separable joint distribution
(`sample_joint_depth_property_copula()`), so both are satisfied by construction rather than left
to chance. This default flip followed multi-AOI real-data validation (a second AOI beyond this
vignette's own Amador County data, plus a full-AOI performance check finding no meaningful
overhead) and closing every other blocking gap identified before the switch - see
`VERTICAL_CORRELATION_IMPROVEMENT_PLAN.md`'s own decision-tracking sections for the full record.

To A/B the two methods fairly, this section reuses the SAME real fitted GP models from Step 3
(`clay_gp_model`/`sand_gp_model`, for `xv_group`) but needs a different starting fixture than Step
6's `clay_mat`/`sand_mat`: those matrices deliberately replicate one shared per-simulation draw
across every depth (`byrow = TRUE`, no fresh noise per depth), which is perfect for demonstrating
the reference-quantile trick but is *already* perfectly depth-correlated by construction - not a
useful "before" state for showing whether a method actually *adds* vertical correlation. Here,
each depth instead gets its own fresh, independent draw (still with the same `target_rho`
clay-sand property correlation established in Step 6):

```{r}
set.seed(4041)
vc_n_sims <- 400
vc_depths <- c(0, 20, 40, 60, 80)

vc_clay_mat <- matrix(NA_real_, nrow = length(vc_depths), ncol = vc_n_sims)
vc_sand_mat <- matrix(NA_real_, nrow = length(vc_depths), ncol = vc_n_sims)
for (i in seq_along(vc_depths)) {
  z_c <- rnorm(vc_n_sims)
  z_s <- target_rho * z_c + sqrt(1 - target_rho^2) * rnorm(vc_n_sims)
  vc_clay_mat[i, ] <- clay_base + 6 * z_c
  vc_sand_mat[i, ] <- sand_base + 8 * z_s
}

# Clay's own correlation between the surface and deepest demo depths - near zero by construction,
# unlike Step 6's fixture (which starts at ~1.0 because every depth is a literal copy).
cor(vc_clay_mat[1, ], vc_clay_mat[nrow(vc_clay_mat), ])
```

Reshape to the long format `apply_gp_depth_trends()` consumes (same `expand.grid()`/column-major
flatten pattern Step 8 already used), and build the `gp_predictions`/`gp_models` lists keyed by
the same property names this cokey data uses:

```{r}
vc_cokey_data <- expand.grid(hzdept_r = vc_depths, simulation_number = seq_len(vc_n_sims))
vc_cokey_data$claytotal <- as.vector(vc_clay_mat)
vc_cokey_data$sandtotal <- as.vector(vc_sand_mat)
vc_cokey_data$cokey <- "vc_demo_cokey"

vc_gp_predictions <- list(
  claytotal = predict_gp_depth_trends(clay_gp_model, vc_depths),
  sandtotal = predict_gp_depth_trends(sand_gp_model, vc_depths)
)
vc_gp_models <- list(claytotal = clay_gp_model, sandtotal = sand_gp_model)
```

Run `apply_gp_depth_trends()` twice on the identical input, explicitly requesting each method (so
this comparison stays valid regardless of which one is the package's current default) - the
original `"gp_quantile_retrofit"` algorithm, still fully supported as an opt-out, and
`"joint_copula"` (passing the real `gp_models` through so the depth kernel's length-scale is
derived from these actual fitted models, per `extract_depth_length_scale()`, rather than falling
back to a generic full-depth-range default):

```{r}
retrofit_result <- apply_gp_depth_trends(
  vc_cokey_data, vc_gp_predictions, c("claytotal", "sandtotal"),
  config = list(monte_carlo = list(vertical_correlation_method = "gp_quantile_retrofit"))
)

joint_result <- apply_gp_depth_trends(
  vc_cokey_data, vc_gp_predictions, c("claytotal", "sandtotal"),
  config = list(monte_carlo = list(vertical_correlation_method = "joint_copula")),
  gp_models = vc_gp_models
)
```

`validate_joint_correlation_structure()` - the same acceptance-test helper
`VERTICAL_CORRELATION_IMPROVEMENT_PLAN.md`'s Phase 0 built for exactly this comparison - reports
both halves of the joint structure (property correlation at every depth, depth correlation for
every property) against a known target, for each method's output:

```{r}
sim_nums <- sort(unique(vc_cokey_data$simulation_number))
retrofit_matrices <- convert_to_property_matrices(retrofit_result, c("claytotal", "sandtotal"), vc_depths, sim_nums)
joint_matrices <- convert_to_property_matrices(joint_result, c("claytotal", "sandtotal"), vc_depths, sim_nums)

target_prop_corr <- matrix(
  c(1, target_rho, target_rho, 1), nrow = 2,
  dimnames = list(c("claytotal", "sandtotal"), c("claytotal", "sandtotal"))
)

retrofit_check <- validate_joint_correlation_structure(retrofit_matrices, target_property_corr = target_prop_corr)
joint_check <- validate_joint_correlation_structure(joint_matrices, target_property_corr = target_prop_corr)

data.frame(
  method = c("gp_quantile_retrofit (production default)", "joint_copula (new, opt-in)"),
  property_corr_max_diff = c(retrofit_check$overall_property_max_diff, joint_check$overall_property_max_diff),
  clay_surface_vs_deepest_corr = c(
    retrofit_check$achieved_depth_correlation$claytotal[1, length(vc_depths)],
    joint_check$achieved_depth_correlation$claytotal[1, length(vc_depths)]
  )
)
```

```{r}
depth_cor_df <- rbind(
  data.frame(
    depth = vc_depths,
    correlation = retrofit_check$achieved_depth_correlation$claytotal[1, ],
    method = "gp_quantile_retrofit\n(production default)"
  ),
  data.frame(
    depth = vc_depths,
    correlation = joint_check$achieved_depth_correlation$claytotal[1, ],
    method = "joint_copula\n(new, opt-in)"
  )
)

ggplot(depth_cor_df, aes(x = depth, y = correlation, color = method)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_color_viridis_d(end = 0.8) +
  theme_minimal() +
  labs(
    title = "Clay content: correlation with the surface depth (0cm), by method",
    subtitle = "Identical starting data (near-zero depth correlation at every depth) fed through both methods",
    x = "Depth (cm)", y = "Correlation with clay at 0cm", color = NULL
  ) +
  theme(legend.position = "bottom")
```

`extract_depth_length_scale(clay_gp_model)` reports `r round(extract_depth_length_scale(clay_gp_model), 1)`
cm for `clay_gp_model`'s own fitted depth trend - a fairly SHORT range, meaning the training data
this GP was fit from shows clay content varying fairly quickly with depth for this soil group. The
`joint_copula` line above tracks that: correlation drops from `1` at the surface to roughly its
`exp(-1) ~= 0.37` "range" value by ~13cm and is essentially gone by 40cm+ (matching an exponential
kernel with that length-scale), the same real, GP-derived smoothness this vignette's other steps
already established for this group.

Both methods keep the clay-sand property correlation close to its `r target_rho` target
(`property_corr_max_diff` small for both) - neither one distorts the correlation structure that was
already there. The depth axis is where they genuinely diverge, and not in the direction a "the old
method under-correlates" story would predict: `gp_quantile_retrofit`'s fixed-per-realization-rank
mechanism pushes clay's own cross-depth correlation to `> 0.99` at every depth tested, regardless
of distance from the surface - a rigidity that has nothing to do with how quickly clay actually
varies with depth in this soil group's own training data. `joint_copula`, by contrast, reproduces a
decay directly tied to `clay_gp_model`'s own fitted length-scale, dropping toward zero within a few
tens of centimeters - the behavior a genuinely physically-motivated vertical correlation model
should have, and the concrete gap `VERTICAL_CORRELATION_IMPROVEMENT_PLAN.md`'s redesign was built
to close.

This is real AOI data (Amador County SSURGO, via the same cached data and fitted GP models used
throughout this vignette) run through both methods with identical synthetic Monte Carlo-style
inputs - offline-reproducible validation of the joint-copula redesign's core claim, without
requiring a live network call. `VERTICAL_CORRELATION_IMPROVEMENT_PLAN.md`'s later phases extended
this validation to a second, independent AOI (finding the same result), confirmed negligible
full-AOI performance overhead, and closed the remaining top-level API gaps - `joint_copula` is now
reachable through `simulate_ssurgo_mapunit_draws()`/`integrate_monte_carlo_with_gp()`'s own
top-level config, not only via `apply_gp_depth_trends()` directly, and is now the package default.
What remains open per that plan's own decision-tracking notes: calibrating the depth kernel's
discontinuity-gating defaults (`distinctness_range`/`min_gate_weight`, still off by default via
`vertical_correlation_gating = FALSE`) against real KSSL/SSURGO profile lag-correlations.

## Where this data came from

The cached data this vignette loads (`inst/extdata/ssurgo_amador.rds`) was produced once by
`data-raw/build_vignette_data.R`, which calls `download_and_prepare_ssurgo()` live against NRCS Soil
Data Access for the AOI used above - the same cached file `getting-started-monte-carlo.Rmd` uses.
Re-run that script to refresh it.
