---
title: "Profile, Component & Depth Simulation with a Real Soil Component"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Profile, Component & Depth Simulation with a Real Soil Component}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6, fig.height = 4)
```

## Overview

soilSIM can simulate two related kinds of variability for a real SSURGO map unit: how much of a map
unit a component actually occupies (component composition), and how variable a component's horizon
depths and thicknesses are in the field (depth/thickness variability, informed by Official Series
Description boundary-distinctness data). This vignette demonstrates both, using a real map unit from
the same Amador-area AOI as the "Getting Started" vignette. See
`soilSIM/docs/06_profile_component_depth_simulation.md` for the full function-level reference.

```{r setup}
library(soilSIM)
library(ggplot2)
```

## The real component

`get_aws_data_by_mukey()` queries NRCS Soil Data Access live for a given map unit key. This
vignette's cached data was produced by that real call (see "Where this data came from" below) for
map unit **462255** within the Amador-area AOI, which resolves to a single real component:

```{r}
depth_sim_mukey <- readRDS(system.file("extdata", "depth_sim_mukey_data.rds", package = "soilSIM"))
mu_data <- depth_sim_mukey$data
mu_data[, c("cokey", "compname", "comppct_l", "comppct_r", "comppct_h",
            "hzname", "hzdept_r", "hzdepb_r", "sandtotal_r", "claytotal_r")]
```

This is a real component named **"`r unique(mu_data$compname)`"** (`cokey`
`r unique(mu_data$cokey)`), with 3 horizons and a representative component percentage of
`r unique(mu_data$comppct_r)`% (ranging `r unique(mu_data$comppct_l)`-`r unique(mu_data$comppct_h)`%).

## Step 1: Simulate component composition

`sim_component_comp()` draws `n_simulations` triangular samples from each component's
low/representative/high percentage and derives a single `sim_comppct` value - the number of
Monte Carlo profile realizations to simulate for that component:

```{r}
component_sim <- sim_component_comp(mu_data, n_simulations = 25)
component_sim
```

With this component's real `r unique(mu_data$comppct_l)`/`r unique(mu_data$comppct_r)`/`r unique(mu_data$comppct_h)`
percentage triplet, 25 triangular draws collapse to `sim_comppct` = `r component_sim$sim_comppct` -
the number of profile replicates simulated in the next step.

## Step 2: Simulate horizon depth and thickness variability

`simulate_profile_depths_by_mukey()` orchestrates the full depth-simulation pipeline for a real
mukey: it queries SSURGO, derives `sim_comppct` via `sim_component_comp()` and joins it onto every
horizon (internally - no manual join needed by the caller), fetches real Official Series
Description boundary-distinctness data for the component name, and perturbs horizon thickness and
boundary depths accordingly:

```{r, eval = FALSE}
# The real call this vignette's cached data came from:
simulated_profiles <- simulate_profile_depths_by_mukey("462255", n_simulations = 25, seed = 123)
```

```{r}
simulated_profiles <- readRDS(system.file("extdata", "depth_sim_profiles_amador.rds", package = "soilSIM"))
simulated_profiles
```

`r length(simulated_profiles)` simulated profile realizations were produced - matching the
`sim_comppct` value derived in Step 1. Each realization independently perturbs the three horizons'
boundary depths:

```{r}
horizons_df <- aqp::horizons(simulated_profiles)
horizons_df[horizons_df$hzname == "H2", c("id", "hzname", "hzdept_r", "hzdepb_r", "thickness_sd")]
```

The second horizon's simulated bottom depth varies across realizations (thickness standard
deviation `r round(unique(horizons_df$thickness_sd[horizons_df$hzname == "H2"]), 2)` cm) - real,
component-specific variability rather than a fixed SSURGO representative value. Plotting the
simulated bottom-depth spread per horizon:

```{r}
ggplot(horizons_df, aes(x = hzname, y = hzdepb_r, fill = hzname)) +
  geom_boxplot() +
  scale_fill_viridis_d(guide = "none") +
  labs(title = paste0(unique(mu_data$compname), ": simulated horizon bottom depths"),
       x = "Horizon", y = "Bottom depth (cm)") +
  theme_minimal()
```

## Visualizing the simulated profiles directly

Boxplots summarize the spread of bottom depths per horizon, but they don't show what any single
simulated profile actually looks like, or how boundary perturbations compound down a profile.
`aqp::plotSPC()` is the standard way to visualize a `SoilProfileCollection`: horizontal soil-profile
"sketches," stacked side by side, one per realization, color-coded by horizon designation. Here all
`r length(simulated_profiles)` simulated realizations are shown together, making the real perturbed
depth boundaries - and how they vary realization to realization - visible at a glance:

```{r, fig.width = 8, fig.height = 5}
par(mar = c(1, 1, 3, 1))
aqp::plotSPC(simulated_profiles, color = "hzname", width = 0.3,
             name.style = "center-center", cex.names = 0.6, cex.id = 0.6,
             main = paste0(unique(mu_data$compname), ": simulated profile realizations"))
```

Beyond bottom-depth position, the actual simulated *thickness* of each horizon
(`hzdepb_r - hzdept_r` within each realization) is another useful view of the same variability,
directly comparable to the `thickness_sd` values driving the simulation:

```{r}
horizons_df$thickness <- horizons_df$hzdepb_r - horizons_df$hzdept_r

ggplot(horizons_df, aes(x = hzname, y = thickness, fill = hzname)) +
  geom_violin(trim = FALSE) +
  geom_boxplot(width = 0.1, fill = "white", alpha = 0.6, outlier.shape = NA) +
  scale_fill_viridis_d(guide = "none") +
  labs(title = paste0(unique(mu_data$compname), ": simulated horizon thickness variability"),
       x = "Horizon", y = "Thickness (cm)") +
  theme_minimal()
```

## Known limitation this vignette's data illustrates

The third horizon (H3) in this real component is missing its `sandtotal_h`/`claytotal_h` (upper
bound) values - a genuine SSURGO data gap, not a simulation artifact. soilSIM's
`evaluate_simulated_depths()` QC function (see `soilSIM/docs/06_profile_component_depth_simulation.md`)
flags exactly this kind of gap: it can only assess whether a simulated depth falls outside a
horizon's original low/high bounds when those bounds are actually present in the source data.

## Where this data came from

The cached data this vignette loads (`inst/extdata/depth_sim_mukey_data.rds`,
`inst/extdata/depth_sim_profiles_amador.rds`) was produced once by `data-raw/build_vignette_data.R`,
which calls `process_aoi_and_get_mukeys_working()`, `get_aws_data_by_mukey()`, and
`simulate_profile_depths_by_mukey()` live against NRCS Soil Data Access and the Official Series
Description database. Re-run that script to refresh it (the real mukey it resolves to may change
if SSURGO data for this AOI is updated).
