---
title: "Multi-Source Raster Fusion: SSURGO x SOLUS100"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multi-Source Raster Fusion: SSURGO x SOLUS100}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6, fig.height = 5)
```

## Overview

This vignette is the clearest example of soilSIM's core architecture: a generic, data-source-agnostic
raster fusion **core** (`fuse_property_adaptive()`, built on the same Bayesian-updating primitives
as the scalar `core-fusion.R` toolkit) combined with two concrete data-source **adapters** -
SSURGO (supplying the prior) and SOLUS100 (supplying the likelihood) - over a shared, real area of
interest (AOI). The architecture is designed so future data sources (e.g. HWSD, SoilGrids) could
plug in as new adapters supplying either side, without changing the fusion core. See
the "Multi-Source Raster Fusion Pipeline" architecture article for the full function-level reference.

```{r setup}
library(soilSIM)
library(ggplot2) # only for the non-spatial prior/likelihood/posterior comparison chart below
```

## The real AOI

A ~2 km x 2 km area of interest in the Salinas Valley, California - chosen because it has confirmed
real SOLUS100 coverage:

```{r}
salinas_wkt <- "POLYGON((-121.66 36.60, -121.64 36.60, -121.64 36.62, -121.66 36.62, -121.66 36.60))"
aoi <- terra::vect(salinas_wkt, crs = "epsg:4326")
aoi <- terra::project(aoi, "epsg:5070")
```

## Step 1: Fuse a single property (clay content)

`run_fusion()` is the top-level orchestrator: it fetches (and disk-caches) SSURGO percentile
rasters as the prior, SOLUS100 percentile rasters as the likelihood, aligns the two onto one shared
grid, and fuses them via `fuse_property_adaptive()`, which adaptively routes to a
closed-form or general grid-KDE fusion depending on the property's distributional shape. By default
(`resampling = "down"`) the SSURGO prior is resampled (bilinear) onto SOLUS100's coarser 100 m grid,
so the fused output sits at SOLUS100's native resolution. Setting `resampling = "up"` instead
resamples SOLUS100 (nearest-neighbor) onto SSURGO's finer native grid, so the fused output preserves
SSURGO's real map-unit boundary detail instead of blurring it away - see the comparison at the end
of this step for what that tradeoff actually buys:

```{r, eval = FALSE}
# The real call this vignette's cached data came from:
property_config <- list(id = "clay", solus_variable = "claytotal", dist = "normal")
fusion_clay <- run_fusion(aoi, property_config, top_depth = 0, bottom_depth = 5)
```

```{r}
fusion_clay <- unwrap_nested_rasters(
  readRDS(system.file("extdata", "fusion_clay_salinas.rds", package = "soilSIM"))
)
fusion_clay$dist
fusion_clay$route
names(fusion_clay$posterior)
```

The fused posterior for a `dist = "normal"` property routed through
`r fusion_clay$route` is a pair of rasters - `mu` (posterior mean) and `sigma` (posterior standard
deviation) - one value per grid cell, at SOLUS100's native 100 m resolution:

```{r}
fusion_clay$posterior$mu
terra::plot(fusion_clay$posterior$mu, main = "Fused posterior mean clay content (%)",
            mar = c(3.1, 3.1, 5.1, 7.1))
```

```{r}
terra::plot(fusion_clay$posterior$sigma, main = "Fused posterior clay uncertainty (SD)",
            mar = c(3.1, 3.1, 5.1, 7.1))
```

### Comparing the prior, likelihood, and fused posterior

The plots above show only the fusion *output*. To see what fusion actually did, we need to look at
its two *inputs* side by side with the result. `fusion_clay$prior$values` and
`fusion_clay$likelihood$values` are each a list of percentile rasters (SSURGO's and SOLUS100's,
respectively) rather than a single raster - we use each source's median (`P50`) as its
representative value:

```{r}
names(fusion_clay$prior$values)      # SSURGO percentiles
names(fusion_clay$likelihood$values) # SOLUS100 percentiles

prior_r <- fusion_clay$prior$values[["P50"]]
likelihood_r <- fusion_clay$likelihood$values[["P50"]]
posterior_r <- fusion_clay$posterior$mu
```

`run_fusion()` already resampled the SSURGO prior onto SOLUS100's grid before fusing, so all
three rasters share the same extent/resolution and combine directly with `c()` - no separate
`par(mfrow = ...)` panels needed. We use one shared color scale (`range`) across all three panels so
that color differences reflect real differences in clay content, not each panel's own auto-rescaling:

```{r, fig.width = 9, fig.height = 3.5}
shared_range <- range(
  terra::minmax(prior_r), terra::minmax(likelihood_r), terra::minmax(posterior_r)
)

compare_stack <- c(prior_r, likelihood_r, posterior_r)
names(compare_stack) <- c("SSURGO prior", "SOLUS likelihood", "Fused posterior")
terra::plot(
  compare_stack,
  main = c("SSURGO prior (median)", "SOLUS likelihood (median)", "Fused posterior mean"),
  col = grDevices::hcl.colors(50, "viridis"),
  range = shared_range,
  nc = 3,
  mar = c(3.1, 3.1, 5.1, 7.1)
)
```

The SSURGO prior is sparse and blocky (map-unit polygons rasterized onto the grid, with `NA` where no
delineation lines up with a cell) while the SOLUS100 likelihood is a smooth, fully-populated
100 m grid - the fused posterior inherits the prior's spatial support (cells stay `NA` unless both
inputs contribute) but its clay values are pulled toward the SOLUS100 pattern.

### Choosing the fusion grid: `resampling = "down"` vs `"up"`

Everything above ran with the default `resampling = "down"`: the SSURGO prior (native ~30 m grid,
from `mukey.wcs()`) is resampled *down* onto SOLUS100's coarser 100 m grid before fusing, so the
fused output above sits at SOLUS100's resolution and SSURGO's map-unit boundaries get blurred into
100 m cells in the process. Passing `resampling = "up"` instead resamples SOLUS100 *up* onto
SSURGO's finer native grid (nearest-neighbor, not interpolation), so the fused output preserves
SSURGO's real polygon boundaries instead:

```{r, eval = FALSE}
# The real call this section's cached "up" data came from - reuses the same live SSURGO/SOLUS
# fetch as fusion_clay above (only the alignment step differs), via run_fusion()'s disk cache:
fusion_clay_up <- run_fusion(aoi, property_config, top_depth = 0, bottom_depth = 5,
                              resampling = "up")
```

```{r}
fusion_clay_up <- unwrap_nested_rasters(
  readRDS(system.file("extdata", "fusion_clay_salinas_up.rds", package = "soilSIM"))
)
```

This is **not** free extra resolution, and the real numbers below show why - side by side with the
`"down"` fusion computed earlier from the exact same underlying SSURGO/SOLUS100 fetch:

```{r}
grid_stats <- function(r, label) {
  data.frame(
    resampling = label,
    rows = terra::nrow(r), cols = terra::ncol(r), n_cells = terra::ncell(r),
    res_m = terra::res(r)[1]
  )
}
rbind(
  grid_stats(fusion_clay$posterior$mu, "down (SOLUS100's grid)"),
  grid_stats(fusion_clay_up$posterior$mu, "up (SSURGO's grid)")
)
```

- **Cell count.** `"up"` fuses over about 11x more cells than `"down"` for this AOI - not a free
  resolution gain, just a real, non-trivial compute-cost increase for the general grid-KDE route in
  particular.
- **Which side is fabricated.** `"down"`'s bilinear step blurs SSURGO's real polygon lines into
  100 m averages - a genuine loss of SSURGO's own information. `"up"`'s nearest-neighbor step
  ("block replication") does not interpolate SOLUS100 at all - every fine cell gets its enclosing
  100 m cell's exact published value, so nothing about SOLUS100 is invented, but nothing about its
  *resolution* improves either: it's still a 100 m product, just tiled onto smaller cells.
  Deliberately nearest-neighbor rather than bilinear here, for the same reason the mukey raster is
  always aligned with `method = "near"` - interpolating between two independent SOLUS100 cell
  predictions would imply a smooth underlying gradient the model never actually asserted.
- **Posterior sigma barely moves, as it should.** Finer cells don't mean a more "confident" fused
  estimate just because there are more of them:

```{r}
data.frame(
  resampling = c("down", "up"),
  mean_posterior_sigma = c(
    terra::global(fusion_clay$posterior$sigma, "mean", na.rm = TRUE)[1, 1],
    terra::global(fusion_clay_up$posterior$sigma, "mean", na.rm = TRUE)[1, 1]
  )
)
```

  The two are within 0.4% of each other - `"up"`'s finer grid is not silently manufacturing extra
  precision, exactly because SOLUS100's likelihood value is locally constant across every SSURGO
  cell inside one 100 m footprint; the real uncertainty (how much SSURGO and SOLUS100 disagree) is
  essentially unchanged by the choice of output grid.

- **Reversibility.** Aggregating the `"up"`-fused posterior mean back onto the `"down"` grid
  (area-weighted average) should reproduce the direct `"down"` fusion if the two are really the same
  underlying answer at different resolutions:

```{r, fig.width = 7, fig.height = 5}
agg_up_to_down <- terra::resample(fusion_clay_up$posterior$mu, fusion_clay$posterior$mu, method = "average")
reversibility_diff <- agg_up_to_down - fusion_clay$posterior$mu
terra::global(abs(reversibility_diff), c("mean", "max"), na.rm = TRUE)

terra::plot(
  reversibility_diff,
  main = "Reversibility check: (up aggregated to 100 m) minus (direct down fusion), clay %",
  col = grDevices::hcl.colors(50, "RdBu", rev = TRUE),
  mar = c(3.1, 3.1, 5.1, 7.1)
)
```

  The mean absolute difference is under 1 percentage point of clay content, confirming `"up"` and
  `"down"` are substantively the same fused answer at two different resolutions - not two
  different pipelines. The larger local differences that remain concentrate where SSURGO's finer
  prior structure inside one SOLUS100 cell pulls the `"up"` fusion's local Bayesian update away
  from what the same cell's single blurred `"down"` prior value would have produced - exactly the
  map-unit detail `"up"` exists to keep instead of averaging away.

In short: use `"up"` when SSURGO's polygon boundaries themselves are the information you need
preserved (e.g. showing where a fused estimate changes at a map-unit line); keep the default
`"down"` when SOLUS100's smoother, denser signal is what should drive the fused output's spatial
pattern, since `"down"` is markedly cheaper and, as the sigma comparison above confirms, doesn't
trade away any real precision to get there.

The same three rasters, viewed as distributions rather than maps, show what the Bayesian update did
numerically:

```{r, fig.width = 7, fig.height = 4.5}
prior_vals <- terra::values(prior_r, na.rm = TRUE)
likelihood_vals <- terra::values(likelihood_r, na.rm = TRUE)
posterior_vals <- terra::values(posterior_r, na.rm = TRUE)

clay_dist_df <- data.frame(
  value = c(prior_vals, likelihood_vals, posterior_vals),
  source = factor(
    c(rep("SSURGO prior", length(prior_vals)),
      rep("SOLUS likelihood", length(likelihood_vals)),
      rep("Fused posterior", length(posterior_vals))),
    levels = c("SSURGO prior", "SOLUS likelihood", "Fused posterior")
  )
)

ggplot(clay_dist_df, aes(x = value, fill = source)) +
  geom_density(alpha = 0.5, color = NA) +
  scale_fill_viridis_d(name = NULL) +
  labs(
    title = "Clay content distributions: prior, likelihood, and fused posterior",
    x = "Clay content (%)",
    y = "Density"
  ) +
  theme_minimal()
```

In this AOI, the SSURGO prior's mass sits mostly at low clay values with a long thin tail, while the
SOLUS100 likelihood is centered much higher (~15-25%). The fused posterior is a sharp, narrow peak
near the prior's low end - the closed-form normal-normal update (`r fusion_clay$route`) combines
both sources weighted by their relative (un)certainty, and here it lands close to the prior rather
than the likelihood, illustrating that "fusion" is a precision-weighted compromise, not a simple
average of the two inputs.

## Step 1b: Percentile output for risk and uncertainty analysis

The `mu`/`sigma` pair above is a good point estimate, but risk- and cost-assessment workflows often
need more than a mean and a standard deviation - exceedance probabilities, asymmetric credible
intervals, or simply "how uncertain is this cell, really?" Every fusion route also returns a full
set of posterior percentiles alongside the point estimate, precisely for this purpose:

```{r}
names(fusion_clay$posterior$percentiles)
```

Nine percentiles by default (`FUSE_POSTERIOR_DEFAULT_PROBS`), covering a 98% interval (P01/P99), a
90% interval (P05/P95), and an 80% interval (P10/P90) around the median - deliberately richer than
the 5-point set SSURGO/SOLUS happen to supply as *input*, since the posterior has no such external
constraint and tail resolution is exactly what risk-style questions need.

### An uncertainty map: the 90% credible interval width

Subtracting the 5th percentile from the 95th gives a per-cell width - a map of *where* the fused
estimate is trustworthy versus where it's genuinely uncertain, not just a single AOI-wide number:

```{r, fig.width = 7, fig.height = 5}
p05 <- fusion_clay$posterior$percentiles$P5
p95 <- fusion_clay$posterior$percentiles$P95
credible_width <- p95 - p05

terra::plot(
  credible_width,
  main = "90% credible interval width (P95 - P5) for clay content (%)",
  col = grDevices::hcl.colors(50, "inferno", rev = TRUE),
  mar = c(3.1, 3.1, 5.1, 7.1)
)
```

This map is not spatially uniform - width narrows where the SOLUS100 likelihood is more informative
relative to the SSURGO prior for that cell, and widens where the two sources disagree more or where
either is individually less certain. A single `sigma` raster carries the same information as a
symmetric summary, but the percentile pair generalizes to cases where the posterior isn't
symmetric (e.g. the closed-form beta/gamma routes, or the general grid-KDE route on a skewed
input) - `sigma` alone can't distinguish "narrow but skewed" from "wide but symmetric," while a
P05/P95 pair always describes the actual shape.

### A risk-style question: where might clay content be higher than expected?

A land-use decision that only tolerates a small chance of encountering high-clay soil (e.g. for
drainage or foundation planning) cares less about the *mean* clay content than about a
conservative upper bound. The P95 percentile answers "what's the clay content this cell is
unlikely (5% chance) to exceed" directly, without assuming normality:

```{r, fig.width = 7, fig.height = 5}
elevated_clay_risk <- p95 > 25
terra::plot(
  elevated_clay_risk,
  main = "Cells where P95 clay content exceeds 25% (elevated-clay risk)",
  col = c("grey85", "firebrick"),
  mar = c(3.1, 3.1, 5.1, 7.1)
)
```

Cells flagged here could plausibly have clay content above 25% even though their *mean* (`mu`) may
sit comfortably below that threshold - exactly the kind of tail information a point estimate alone
discards, and exactly why the percentile output exists alongside it rather than replacing it.

## Step 2: Fuse a compositional group (clay, sand, silt jointly)

Fusing clay/sand/silt independently can break the sum-to-100 constraint texture fractions must
satisfy. `run_fusion_group()` fuses a whole compositional group jointly, in isometric
log-ratio (ILR) space, guaranteeing every cell's fused clay + sand + silt sums to 100:

```{r, eval = FALSE}
# The real call this vignette's cached data came from:
composition_groups <- list(texture = list(members = c("clay", "sand", "silt")))
property_configs <- list(
  clay = list(id = "clay", solus_variable = "claytotal", composition_group = "texture"),
  sand = list(id = "sand", solus_variable = "sandtotal", composition_group = "texture"),
  silt = list(id = "silt", solus_variable = "silttotal", composition_group = "texture")
)
fusion_texture <- run_fusion_group(
  aoi, "texture", composition_groups, property_configs, top_depth = 0, bottom_depth = 5
)
```

```{r}
fusion_texture <- unwrap_nested_rasters(
  readRDS(system.file("extdata", "fusion_texture_salinas.rds", package = "soilSIM"))
)
names(fusion_texture)
fusion_texture$clay$dist   # "texture_ilr" for every member - not the per-property normal/beta/gamma dist
```

```{r}
means <- sapply(fusion_texture, function(m) terra::global(m$posterior$value, "mean", na.rm = TRUE)[1, 1])
means
sum(means)
```

Mean fused clay/sand/silt percentages sum to `r round(sum(means), 1)` across the AOI - the joint ILR
fusion's guarantee, verified here on real fused output rather than just asserted.

```{r}
terra::plot(c(fusion_texture$clay$posterior$value,
              fusion_texture$sand$posterior$value,
              fusion_texture$silt$posterior$value),
            main = c("Clay", "Sand", "Silt"),
            mar = c(3.1, 3.1, 5.1, 7.1))
```

Each member also carries its own fraction's posterior percentiles, drawn from a dedicated
raster-native sampler of the fused bivariate-Normal ILR posterior:

```{r}
names(fusion_texture$clay$posterior$percentiles)
```

**Sum-to-100 caveat**: `value` (the point estimate above) sums to exactly 100 by construction - the
whole reason this vignette's joint ILR fusion exists. Per-fraction *percentiles* do **not** share
that guarantee, and this isn't a bug:

```{r}
p95_sum <- Reduce(`+`, lapply(fusion_texture, function(m) m$posterior$percentiles$P95))
terra::global(p95_sum, fun = c("min", "max"), na.rm = TRUE)
```

The P95 percentiles across clay/sand/silt sum well above 100 in this AOI - percentiles of
correlated marginals don't add linearly the way means do (a cell can simultaneously see a
high-tail clay draw and a high-tail sand draw from different regions of the same joint
distribution). Use `value` when a hard sum-to-100 composition is required; use `percentiles` for
per-fraction uncertainty, not as a second composition that should also sum to 100.

## Under the hood: scalar Bayesian updating

Both fusion routes above ultimately reduce to elementwise arithmetic on `core-fusion.R`'s
scalar fusion functions (`fuse_normal_normal()`, `fuse_beta()`/`fuse_gamma()`,
`update_prior()`'s general grid-KDE fusion) - they already work unchanged on `SpatRaster` inputs
via `terra`'s operator overloading, so the raster-native code in `core-fusion.R` reuses them
directly rather than reimplementing the same math per-cell. See
the "Bayesian Updating" architecture article for these building blocks used directly on scalar
prior/likelihood pairs, independent of any raster or AOI.

## Where this data came from

The cached data this vignette loads (`inst/extdata/fusion_clay_salinas.rds`,
`inst/extdata/fusion_texture_salinas.rds`) was produced once by `data-raw/build_vignette_data.R`,
which calls `run_fusion()`/`run_fusion_group()` live against NRCS Soil Data Access
(SSURGO) and `soilDB::fetchSOLUS()` (SOLUS100) for the AOI above. Re-run that script to refresh it.
Producing this vignette's data surfaced and fixed a real caching bug in
`run_fusion_group()` (it wasn't wrapping/unwrapping `SpatRaster`s around its disk cache,
silently corrupting the cached texture-group result on a second read) - fixed in `R/core-fusion.R`.
