---
title: "Per-Pixel Fused Ensemble & Available Water Capacity"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Per-Pixel Fused Ensemble & Available Water Capacity}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6, fig.height = 5)
```

## Overview

The `raster-fusion-ssurgo-solus` vignette shows how to fuse **one** SSURGO-derived prior against a
**one** SOLUS100 likelihood into a per-cell posterior for a single property. This vignette takes the
next step: it wires that fused posterior back into the tabular Monte Carlo framework as a
**per-pixel joint ensemble**, so that a derived quantity - here plant-available water capacity (AWC)
- can be computed per realization and summarised into a genuine per-pixel probability distribution.

The bridge is a **rank-preserving re-marginalization** of a single joint Monte Carlo ensemble, not
an independent per-property resample. Three functions:

| Function | Role |
|---|---|
| `extract_mukey_joint_ensemble()` | one joint SSURGO Monte Carlo, retained per map unit, row-aligned across depth windows |
| `remarginalize_ensemble_to_posterior()` | transform each realization's marginals to the SOLUS-fused posterior, per pixel, keeping the ensemble's rank copula |
| `remarginalized_awc()` | AWC per realization from the re-marginalized inputs, then per-pixel percentiles |

AWC needs special handling: **SOLUS100 publishes no water-retention variable**, so field capacity
and wilting point cannot be fused directly. Instead the fusable drivers - sand, silt, clay, bulk
density, organic carbon - are fused and re-marginalized, and water retention is derived from them
per (pixel, realization) via the Saxton-Rawls pedotransfer function.

See `docs/the "Multi-Source Raster Fusion Pipeline" architecture article` -> "Statistical structure of the per-pixel
bridge" for the full treatment of what the resulting uncertainty band does and does not capture;
this vignette walks the mechanics with real data.

```{r setup}
library(soilSIM)
```

## The AOI

The same ~2 km x 2 km Salinas Valley area of interest as the `raster-fusion-ssurgo-solus` vignette,
chosen for confirmed real SOLUS100 coverage.

```{r}
salinas_wkt <- "POLYGON((-121.66 36.60, -121.64 36.60, -121.64 36.62, -121.66 36.62, -121.66 36.60))"
aoi <- terra::vect(salinas_wkt, crs = "epsg:4326")
aoi <- terra::project(aoi, "epsg:5070")
```

## Step 1: one joint SSURGO Monte Carlo ensemble

`extract_mukey_joint_ensemble()` runs the SSURGO Monte Carlo **once** for the whole AOI over the
requested depth windows. Internally, `simulate_cokey_generalized()` draws every property jointly
under the KSSL reference correlation matrices, the joint-copula vertical-correlation step runs down
each simulated profile, and the result is aggregated to each depth window **row-aligned**: the same
row index in the `0-5` and `5-15` matrices is the *same simulated profile*.

```{r, eval = FALSE}
# The real call this vignette's cached data came from:
windows <- list(c(0, 5), c(5, 15), c(15, 30))
ens <- extract_mukey_joint_ensemble(aoi, windows)
```

```{r}
ens <- readRDS(system.file("extdata", "perpixel_ensemble_salinas.rds", package = "soilSIM"))
ens$mukey_raster <- terra::unwrap(ens$mukey_raster)   # cached wrapped; see "Where this data came from"

ens$window_names
ens$properties
length(ens$by_mukey)                                  # one entry per map unit
```

Each map unit's entry is a set of `[realization x property]` matrices, one per depth window:

```{r}
round(head(ens$by_mukey[[1]]$windows[["0-5"]], 3), 1)
```

Rows are realizations (each a simulated profile draw), columns are properties drawn *jointly*.
Because the windows are row-aligned, realization `r`'s clay content varies with depth but is still
"realization `r`" at every depth - the vertical link that keeps a profile coherent:

```{r}
vapply(ens$window_names,
       function(w) ens$by_mukey[[1]]$windows[[w]][1, "clay_total"],
       numeric(1))
```

## Step 2: fuse every Saxton-Rawls input against SOLUS in one pass

`run_stage1_fusion_multi()` produces a per-cell SSURGO x SOLUS posterior for **many** properties
across **many** depth windows, running the (expensive) SSURGO Monte Carlo simulation just once and
reusing that one draw set for every property/window. Calling the single-property
`run_stage1_fusion()` in a loop instead would re-run the full simulation for each of the
`5 inputs x 3 windows = 15` combinations, each discarding all but one of the ~9 jointly simulated
properties. `dist = "auto"` lets the fusion core pick the distribution family per property.

Because every leaf comes from the *same* simulation, the sand/silt/clay posteriors share an exact
`NA` mask (they are one joint texture draw), and each property's coverage is a stable, reproducible
consequence of that one run rather than of five independent ones. Texture and bulk density can
still differ: a map unit whose components have complete bulk-density data but no texture triplet
contributes `db` but drops out of the texture columns entirely.

```{r, eval = FALSE}
# The real call this vignette's cached data came from - one simulation for all 5 x 3 leaves:
cfg <- function(id, solus_var) list(id = id, solus_variable = solus_var, dist = "auto")
sr_inputs <- c(sand_total = "sandtotal", silt_total = "silttotal", clay_total = "claytotal",
               db = "dbovendry", soc = "soc")
property_configs <- setNames(
  lapply(names(sr_inputs), function(nm) cfg(nm, sr_inputs[[nm]])),
  names(sr_inputs)
)
post <- run_stage1_fusion_multi(aoi, property_configs, windows, simplify = TRUE)
```

```{r}
post <- unwrap_nested_rasters(
  readRDS(system.file("extdata", "perpixel_posteriors_salinas.rds", package = "soilSIM"))
)
names(post)
names(post$clay_total)
names(post$clay_total[["0-5"]]$percentiles)            # nine percentile-value rasters
```

```{r}
terra::plot(post$clay_total[["0-5"]]$percentiles$P50,
            main = "SOLUS-fused clay content, P50, 0-5 cm (%)")
```

## Step 3: re-marginalize the ensemble onto the grid

`remarginalize_ensemble_to_posterior()` maps every retained source realization through
`F_prior` (the map unit's empirical CDF) then `F_posterior^-1` (the pixel's fused posterior):
`v_new = F_posterior^-1(F_prior(v_r))`. Crucially, **one realization index is used for every
property and every window**, so each realization keeps its rank position everywhere - the
transformed ensemble has the source's cross-property *and* cross-depth **rank** correlations, only
the marginals become the fused posteriors.

```{r}
rm <- remarginalize_ensemble_to_posterior(ens, post, n_out = 150, summarize = FALSE)
rm$n_kept
names(rm$ensemble)
terra::nlyr(rm$ensemble$clay_total[["0-5"]])           # one layer per retained realization
```

### The rank copula is preserved

For any pixel, the transformed realizations of two properties keep exactly the Spearman correlation
the source SSURGO ensemble has for that pixel's map unit. Sand and clay are strongly
(negatively) correlated - they are compositional parts of the same texture triangle:

```{r}
w    <- "0-5"
grid <- rm$ensemble$clay_total[[w]][[1]]

# per-cell numeric map unit code on the transformed grid
mkn <- terra::resample(ens$mukey_raster, grid, method = "near")
levels(mkn) <- NULL
mk_cell   <- as.character(terra::values(mkn)[, 1])
grid_vals <- terra::values(grid)[, 1]

# a map unit that has complete texture data AND at least one pixel in this AOI
pick <- NULL
for (m in names(ens$by_mukey)) {
  cm <- which(mk_cell == m & is.finite(grid_vals))
  if (!length(cm)) next
  if (anyNA(ens$by_mukey[[m]]$windows[[w]][seq_len(rm$n_kept), c("sand_total", "clay_total")])) next
  pick <- list(mk = m, cell = cm[1]); break
}

src     <- ens$by_mukey[[pick$mk]]$windows[[w]][seq_len(rm$n_kept), ]
tr_sand <- as.numeric(terra::values(rm$ensemble$sand_total[[w]])[pick$cell, ])
tr_clay <- as.numeric(terra::values(rm$ensemble$clay_total[[w]])[pick$cell, ])

round(c(
  source      = stats::cor(src[, "sand_total"], src[, "clay_total"], method = "spearman"),
  transformed = stats::cor(tr_sand,             tr_clay,             method = "spearman")
), 3)
```

The two values match: the re-marginalization changed each property's marginal distribution to the
fused posterior but left every realization's *rank* essentially untouched (they can differ only in
the 3rd-4th decimal, where a tie forms because the fused posterior's tail is clamped flat).

## Step 4: per-pixel available water capacity

`remarginalized_awc()` (default `method = "saxton_rawls"`) takes, for each realization / pixel /
window, the jointly-consistent `(sand, clay, silt, db, om)` values, derives field capacity and
wilting point via `saxton_rawls_raster()`, and accumulates
`AWC = sum_windows (fc - wp)/100 * thickness` for that realization. `terra::quantile()` over the
retained realizations then gives the per-pixel AWC distribution.

```{r}
awc <- remarginalized_awc(ens, post, n_out = 150)
awc$n_kept
awc$windows_used
names(awc$awc_cm)
```

```{r}
terra::plot(awc$awc_cm$P50,
            main = "Median plant-available water capacity, 0-30 cm (cm)",
            col = grDevices::hcl.colors(50, "Blues 3", rev = TRUE))
```

```{r}
terra::plot(awc$awc_cm$P95 - awc$awc_cm$P5,
            main = "90% credible-interval width for AWC (cm)",
            col = grDevices::hcl.colors(50, "inferno", rev = TRUE))
```

`awc_cm$P95 - awc_cm$P5` is a genuine 90% credible interval because AWC is computed per source
realization from vertically- and cross-property-consistent inputs, then combined per pixel.

## Optional: bedrock/restriction-depth-aware AWC

SOLUS100 predicts every property at every depth even where bedrock is shallower - nothing above
truncates the 30 cm profile at an actual restriction, so a shallow-soil pixel can be credited with
water storage below bedrock. `remarginalized_awc()`'s optional `restriction_depth` argument fixes
this: pass a restriction-depth raster (typically `fetch_solus_restriction_depth()`, which defaults
to SOLUS's `anylithicdpt` - depth to bedrock, trained specifically on lithic/paralithic restriction
records - and is already right-censoring-guarded at 201 cm) and each window's thickness
contribution becomes the per-pixel *effective* thickness instead of its nominal one.

This step needs a live SOLUS100 fetch, so it is not part of this vignette's cached pipeline - shown
for reference, not executed here. `restriction_depth` is opt-in (`NULL` by default preserves every
result above exactly):

```{r, eval = FALSE}
restriction_depth <- fetch_solus_restriction_depth(aoi)  # default variable = "anylithicdpt"
awc_truncated <- remarginalized_awc(ens, post, n_out = 150, restriction_depth = restriction_depth)
terra::plot(awc_truncated$awc_cm$P50,
            main = "Median AWC, 0-30 cm, truncated at bedrock (cm)",
            col = grDevices::hcl.colors(50, "Blues 3", rev = TRUE))
```

## Reading the uncertainty band

| Aspect | Treatment |
|---|---|
| **Marginal** uncertainty of each input property | the full SSURGO x SOLUS fused posterior spread per pixel, propagated through Saxton-Rawls - **captured** |
| Cross-property & cross-depth **dependence** | the SSURGO tabular copula (KSSL + joint-copula vertical correlation), **held fixed** - SOLUS carries no joint information, so nothing is re-estimated |
| Correlation type preserved | **rank (Spearman)**, not linear (Pearson) |
| Pedotransfer-function error | **not propagated** - Saxton-Rawls is applied deterministically per realization |
| Sub-map-unit variation in the **dependence** structure | none - every pixel in a map unit shares the source ensemble; only the fused marginals vary pixel to pixel |
| Texture compositional closure | independent re-marginalization can leave `sand + silt + clay != 100`; `saxton_rawls_raster()` renormalizes to 100 when off by more than 5 |

In short, the per-pixel AWC distribution reflects **SOLUS-fused marginal uncertainty threaded
through the SSURGO tabular copula**, not a fully re-estimated joint posterior and not
pedotransfer-model error.

## Per-pixel vs. map-unit-collapsed

A cheaper alternative would collapse the fused posterior to one distribution per map unit (via
`zonal_distribution_from_posterior()`) and feed that back through the tabular Monte Carlo. How much
does that lose? Compare the within-map-unit spread of the per-pixel AWC to the between-map-unit
spread:

```{r}
p50 <- awc$awc_cm$P50
mkr <- terra::resample(ens$mukey_raster, p50, method = "near")
within  <- terra::zonal(p50, mkr, fun = "sd",   na.rm = TRUE)
between <- terra::zonal(p50, mkr, fun = "mean", na.rm = TRUE)

c(mean_within_mapunit_sd = mean(within[[2]], na.rm = TRUE),
  between_mapunit_sd     = stats::sd(between[[2]], na.rm = TRUE))
```

Whether the ratio is negligible is **property-dependent**: benchmarking on a comparable AOI found it
around 0.01 for clay content and pH (a map-unit-collapsed feed loses almost nothing) but around
0.42 for bulk density. Because AWC is strongly bulk-density-sensitive, the per-pixel path is the
right default for it - `zonal_distribution_from_posterior()` is kept internal.

## Where this data came from

The cached objects this vignette loads (`inst/extdata/perpixel_ensemble_salinas.rds`,
`inst/extdata/perpixel_posteriors_salinas.rds`) were produced once by
`data-raw/build_vignette_data.R`, which runs `extract_mukey_joint_ensemble()` and
`run_stage1_fusion_multi()` live against NRCS Soil Data Access and SOLUS100 for the AOI above, trims
each map unit's ensemble to 250 realizations, wraps the rasters via `terra::wrap()` /
`wrap_nested_rasters()`, and saves. The re-marginalization and AWC steps shown above run live in
this vignette - they are deterministic and need no network access.
