---
title: "Raster-Native Fitting & Fusion Internals"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Raster-Native Fitting & Fusion Internals}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6, fig.height = 4)
```

## Overview

`raster-fusion-ssurgo-solus.Rmd` shows the *pipeline* view of soilSIM's multi-source raster fusion:
call `run_stage1_fusion()`/`run_stage1_fusion_group()` and get back a fused posterior. This
vignette opens the hood. `run_stage1_fusion()` is a thin fetch-and-cache wrapper around a handful
of small, composable, raster-native building blocks - distribution fitting functions
(`R/distribution-fitting-raster.R`), a fusion core (`R/raster-fusion.R`), and a disk cache
(`R/raster-cache.R`) - each of which does exactly one thing to a `terra::SpatRaster` (or list of
them) and can be called directly. Understanding these individually is what makes the pipeline's
behavior (which route it took, why a family was resolved a certain way, why two sources' percentiles
had to be realigned before fusing) legible rather than a black box. See
`soilSIM/docs/09_multi_source_raster_fusion_pipeline.md` for the full function-level reference this
vignette is built from.

```{r setup}
library(soilSIM)
library(terra)
library(ggplot2)
```

## Reusing real data instead of fetching

Fetching SSURGO/SOLUS100 live requires network access and takes minutes; this vignette reuses the
same cached Salinas Valley clay-content fusion result `raster-fusion-ssurgo-solus.Rmd` uses, so every
function below runs on real percentile rasters rather than synthetic ones. The real (network) calls
that originally produced this cached object are shown for reference only and not evaluated here:

```{r, eval = FALSE}
# Reference only - not run in this vignette.
salinas_wkt <- "POLYGON((-121.66 36.60, -121.64 36.60, -121.64 36.62, -121.66 36.62, -121.66 36.60))"
aoi <- terra::project(terra::vect(salinas_wkt, crs = "epsg:4326"), "epsg:5070")
property_config <- list(id = "clay", solus_variable = "claytotal", dist = "normal")
fusion_clay <- run_stage1_fusion(aoi, property_config, top_depth = 0, bottom_depth = 5)
```

```{r}
fusion_clay <- unwrap_nested_rasters(
  readRDS(system.file("extdata", "fusion_clay_salinas.rds", package = "soilSIM"))
)

# The two inputs every function below operates on: named lists of percentile-value
# SpatRasters, one list per source, plus the probabilities each percentile represents.
prior_values <- fusion_clay$prior$values        # SSURGO, already resampled onto the SOLUS grid
prior_probs <- fusion_clay$prior$probs
lik_values <- fusion_clay$likelihood$values     # SOLUS100
lik_probs <- fusion_clay$likelihood$probs

names(prior_values); prior_probs
names(lik_values); lik_probs
bounds <- c(0, 100)  # clay content is a percentage
```

This is the "common currency" every function in this vignette speaks: a **named list of
single-layer percentile-value rasters** (e.g. `P05`, `P25`, `P50`, ...) plus a **numeric vector of
the probabilities they represent**, all sharing one grid. Everything downstream - fitting,
quantile evaluation, fusion, caching - operates on this shape (or a pair of them, prior and
likelihood).

## 1. Fitting a Normal distribution

`fit_normal_raster()` is the simplest fit in the file: closed-form, exact, and just three raster
arithmetic operations. It takes a low/median/high percentile-value raster triplet and the
probabilities the low/high rasters represent, and returns `mu` (the median raster, used directly
as the mean) and `sigma` (back-solved from the tail spacing):

```{r}
fit_n <- fit_normal_raster(
  p_lo_r = prior_values$P25, p50_r = prior_values$P50, p_hi_r = prior_values$P75,
  p_lo = 0.25, p_hi = 0.75
)
fit_n$mu
```

`sigma` is `(p_hi_r - p_lo_r) / (qnorm(p_hi) - qnorm(p_lo))` - the raw value spread between the two
tail percentiles, divided by how many standard-normal units apart those two probabilities are. This
is a single elementwise expression, so its cost doesn't depend on how many cells the raster has.
`quantile_normal_raster()` then evaluates the fitted distribution at any target probability `q` via
`mu + sigma * qnorm(q)`:

```{r, fig.width = 9, fig.height = 3.5}
q10 <- quantile_normal_raster(fit_n, 0.1)
q90 <- quantile_normal_raster(fit_n, 0.9)

normal_stack <- c(q10, fit_n$mu, q90)
names(normal_stack) <- c("Q10", "mu (=P50)", "Q90")
terra::plot(normal_stack, main = c("Normal Q10", "Normal mean", "Normal Q90"),
            col = grDevices::hcl.colors(50, "viridis"), nc = 3)
```

## 2. Fitting a Beta distribution: method-of-moments vs. Newton-Raphson MLE

Clay content is bounded on `[0, 100]`, which the Normal fit above ignores (its Q10/Q90 rasters can
dip below 0 in low-clay cells). A Beta fit respects those bounds. soilSIM ships two Beta fitters:

- `fit_beta_mom_raster(mean_r, var_r)` - closed-form method-of-moments, APPROXIMATE, and used only
  to seed the Newton-Raphson solver below (not intended as a standalone fit). It takes a
  **rescaled-to-`[0,1]`** mean/variance pair directly, not raw percentile rasters.
- `fit_beta_mle_newton_raster(value_rasters, bounds, n_iter = 15, eps = 1e-6)` - a vectorized
  Newton-Raphson MLE solver that exploits the Beta log-likelihood's closed-form score/Hessian (in
  `digamma()`/`trigamma()`, both vectorized base-R functions) to update every cell's `(alpha,
  beta)` simultaneously per iteration, `n_iter` times. It takes the raw percentile-value rasters and
  the physical `bounds` directly, and internally seeds itself from `fit_beta_mom_raster()`.

```{r}
# Method-of-moments needs mean/var already rescaled to [0,1]; reuse the Normal fit above as a
# cheap mean/variance proxy, rescaled by `bounds`.
mean_r <- fit_n$mu / bounds[2]
var_r <- (fit_n$sigma / bounds[2])^2
fit_beta_mom <- fit_beta_mom_raster(mean_r, var_r)

# The Newton-Raphson MLE fit takes the raw percentile rasters and bounds directly - it does its
# own internal method-of-moments seeding.
fit_beta_mle <- fit_beta_mle_newton_raster(prior_values, bounds)
```

```{r, fig.width = 9, fig.height = 6.5}
beta_params <- c(fit_beta_mom$alpha, fit_beta_mom$beta, fit_beta_mle$alpha, fit_beta_mle$beta)
names(beta_params) <- c("alpha (MoM)", "beta (MoM)", "alpha (Newton MLE)", "beta (Newton MLE)")
terra::plot(beta_params, col = grDevices::hcl.colors(50, "plasma"), nc = 2)
```

The two fits' shape-parameter rasters differ cell-by-cell (MoM only matches the first two moments;
Newton MLE fits all of the percentiles supplied to it), but both feed the same quantile function
shape - `quantile_beta_mom_raster(fit, q)` / `quantile_beta_mle_newton_raster(fit, q)`, each
implemented via `terra::lapp()` calling `stats::qbeta()` per cell, since `qbeta()` has no
raster-native vectorized form of its own:

```{r, fig.width = 9, fig.height = 3.5}
q_mom <- quantile_beta_mom_raster(fit_beta_mom, 0.5)
q_mle <- quantile_beta_mle_newton_raster(fit_beta_mle, 0.5)

median_compare <- c(prior_values$P50, q_mom, q_mle)
names(median_compare) <- c("Raw P50 input", "Beta MoM Q50", "Beta Newton-MLE Q50")
terra::plot(median_compare, col = grDevices::hcl.colors(50, "viridis"), nc = 3)
```

The Newton-MLE fit's own Q50 tracks the raw input P50 raster much more closely than the MoM fit's
does, since MLE was fit against all five of `prior_values`' percentiles rather than just a
mean/variance proxy derived from two of them.

## 3. Fitting a metalog distribution, and its feasibility fallback

The metalog family can match an arbitrary number of percentiles exactly via a linear solve rather
than an optimization - `fit_metalog_linear_raster(value_rasters, probs, bounds, boundedness)`
requires the number of *interior* percentiles (excluding `p = 0`/`p = 1`, where the metalog's logit
transform is undefined) to equal the number of metalog terms, making the fit exactly determined.

Symmetric percentile sets are a real, sharp edge here: fitting all five of `prior_values`'s
percentiles (`0.05/0.25/0.5/0.75/0.95` - symmetric about the median) makes the underlying basis
matrix numerically singular, since two of its basis columns become linearly dependent for a
symmetric probability grid. An asymmetric 4-term subset avoids it:

```{r}
ml_probs <- c(0.05, 0.25, 0.5, 0.75)
ml_values <- prior_values[c("P05", "P25", "P50", "P75")]
fit_ml <- fit_metalog_linear_raster(ml_values, ml_probs, bounds, boundedness = "b")
length(fit_ml$a)   # one coefficient raster per term
fit_ml$term
```

Not every fitted metalog is a *valid* one - a metalog quantile function must be monotonically
increasing (equivalent to non-negative density everywhere), and the exact linear solve above gives
no such guarantee. `check_metalog_feasibility_raster()` probes the fit at a grid of probabilities
and flags cells where consecutive probes decrease:

```{r}
infeasible <- check_metalog_feasibility_raster(fit_ml, bounds, boundedness = "b")
n_infeasible <- sum(terra::values(infeasible), na.rm = TRUE)
n_total <- terra::ncell(infeasible)
cat(sprintf("%d / %d cells (%.0f%%) have an infeasible metalog fit\n",
            n_infeasible, n_total, 100 * n_infeasible / n_total))
terra::plot(infeasible, main = "Infeasible metalog fit (TRUE = non-monotonic)",
            col = c("grey85", "firebrick"))
```

A substantial fraction of cells are infeasible here - not a bug, just a real consequence of fitting
only 4 of the 5 available percentiles to a family with no correction step of its own.
`quantile_metalog_linear_with_fallback()` is the function every real caller should use instead of
`quantile_metalog_linear_raster()` directly: it evaluates the raw metalog quantile function, and for
whichever cells `check_metalog_feasibility_raster()` flagged, blends in
`quantile_linear_cdf_raster()`'s nonparametric interpolation (evaluated against the *full*
percentile set, not just the interior ones used to fit the metalog) instead:

```{r, fig.width = 9, fig.height = 3.5}
fallback_result <- quantile_metalog_linear_with_fallback(
  fit_ml, infeasible, full_value_rasters = prior_values, full_probs = prior_probs,
  q = 0.5, bounds = bounds, boundedness = "b"
)

raw_metalog_q50 <- quantile_metalog_linear_raster(fit_ml, 0.5, bounds, boundedness = "b")
blend_stack <- c(raw_metalog_q50, fallback_result$value, prior_values$P50)
names(blend_stack) <- c("Raw metalog Q50 (no fallback)", "With fallback", "Raw input P50")
terra::plot(blend_stack, col = grDevices::hcl.colors(50, "viridis"), nc = 3)
```

## 4. Fitting a Gamma distribution

`fit_gamma_mom_raster()` is a thin raster wrapper: it computes mean/variance rasters across a list
of percentile-value rasters (the only genuinely raster-specific step, via `Reduce(+, ...)`), then
delegates to `R/bayesian-updating.R`'s existing scalar `moments_to_gamma()`, which is pure
elementwise arithmetic and therefore already works unchanged on `SpatRaster` inputs:

```{r, fig.width = 9, fig.height = 3.5}
fit_gamma <- fit_gamma_mom_raster(prior_values)
gamma_stack <- c(fit_gamma$shape, fit_gamma$rate)
names(gamma_stack) <- c("shape", "rate")
terra::plot(gamma_stack, col = grDevices::hcl.colors(50, "inferno"), nc = 2)
```

## 5. Aligning mismatched percentile grids: `align_percentile_probs()`

SSURGO's prior percentiles (`r paste(prior_probs, collapse = ", ")`) and SOLUS100's likelihood
percentiles (`r paste(lik_probs, collapse = ", ")`) aren't the same probabilities - every fusion
route needs both sides to share one `probs` vector, so passing them through mismatched would
silently mis-fit whichever side's percentiles don't match the assumed labels.

```{r}
aligned <- align_percentile_probs(prior_values, prior_probs, lik_values, lik_probs)
aligned$probs
names(aligned$prior_value_rasters)   # unchanged (SSURGO's probs won - see below)
length(aligned$lik_value_rasters)    # SOLUS reinterpolated onto aligned$probs - unnamed list
```

`align_percentile_probs()` picks whichever side's probabilities span the *narrower* range as the
shared target (`diff(range(prior_probs))` = `r round(diff(range(prior_probs)), 3)` vs.
`diff(range(lik_probs))` = `r round(diff(range(lik_probs)), 3)`, so the SSURGO side's probabilities
win here) and reinterpolates the other side onto it via `quantile_linear_cdf_raster()` - never the
reverse, since interpolating requires the target probability to fall within the source's own range,
and the narrower side is always safely interpolatable from the wider one. The reinterpolated side
comes back as a plain (unnamed) list in `aligned$probs`' order, so pick the layer to compare by
matching its position in `aligned$probs` rather than by name:

```{r, fig.width = 9, fig.height = 3.5}
idx50 <- which(aligned$probs == 0.5)
lik_compare <- c(lik_values$P50, aligned$lik_value_rasters[[idx50]])
names(lik_compare) <- c("SOLUS raw P50", "SOLUS P50 re-expressed at aligned probs")
terra::plot(lik_compare, col = grDevices::hcl.colors(50, "viridis"), nc = 2)
```

Here the shared target happens to already include `0.5` on both sides, so the `P50` layer above is
unchanged by realignment - it's the tail percentiles (`P05`/`P95` vs. `P025`/`P975`) that actually
get reinterpolated.

## 6. Family resolution: `resolve_property_dist()`

When `property_config$dist == "auto"`, `resolve_property_dist()` picks a distribution family from
the AOI's own percentile skew rather than a fixed per-property label - and never overrides an
explicit (non-`"auto"`) `dist`. It's deliberately narrow: it only ever resolves to `"beta"` (if
`bounds` is configured), `"normal"`, or `"lognormal"` - never `"metalog"` or `"gamma"`, since
distinguishing those from a 3-point proxy isn't considered reliable.

```{r}
# With bounds configured, "auto" always resolves to "beta" regardless of skew.
resolve_property_dist(list(dist = "auto", bounds = c(0, 100)), prior_values, prior_probs)

# Without bounds, it falls back to a skew-proxy test: normal if |skew| is small, lognormal if not.
resolve_property_dist(list(dist = "auto"), prior_values, prior_probs)
```

The skew proxy itself is `((high - median) - (median - low)) / (high - low)`, aggregated once
across the whole AOI (default `stats::median()`) rather than per cell - computing it per cell would
create map discontinuities right at the family boundary. An explicit `dist` is always honored
untouched:

```{r}
resolve_property_dist(list(dist = "normal"), prior_values, prior_probs)$dist_source
```

## 7. Adaptive fusion dispatch: `fuse_adaptive()`

`fuse_adaptive()` is the size-adaptive engine `fuse_property_adaptive()` (and therefore
`run_stage1_fusion()`) delegates to for `"normal"`/`"beta"`/`"gamma"` families. It picks the fusion
*route* from the AOI's own cell count rather than requiring the caller to choose: at or below
`threshold_cells` (default 80,000), it uses a fully general per-cell KDE route
(`bayesian_update()` on samples drawn from each side's percentiles); above it, a much faster
closed-form route built directly on the fitting functions from Sections 1-4 above (e.g.
`fit_normal_raster()` + `bayes_update_normal_normal()` for `family = "normal"`).

This AOI is small (`r terra::ncell(aligned$prior_value_rasters[[1]])` cells), so it naturally takes
the general route. Forcing `threshold_cells = 0` shows the alternate closed-form route on the exact
same inputs, with `verbose = TRUE` (the pipeline default is quiet - this is a deliberate log demo):

```{r}
fused_general <- fuse_adaptive(
  aligned$prior_value_rasters, aligned$lik_value_rasters, aligned$probs,
  family = "normal", verbose = TRUE
)
fused_closed_form <- fuse_adaptive(
  aligned$prior_value_rasters, aligned$lik_value_rasters, aligned$probs,
  family = "normal", threshold_cells = 0, verbose = TRUE
)
```

```{r, fig.width = 9, fig.height = 3.5}
route_compare <- c(fused_general$posterior$mu, fused_closed_form$posterior$mu)
names(route_compare) <- c(paste0("General route (", fused_general$route, ")"),
                           paste0("Closed-form route (", fused_closed_form$route, ")"))
terra::plot(route_compare, col = grDevices::hcl.colors(50, "viridis"), nc = 2)
```

Both routes fuse the same prior/likelihood into a `mu`/`sigma` posterior with the same output
contract, but they reach it differently: the general route makes no distributional assumption about
the *fusion* step itself (only about the requested output family), while the closed-form route fits
both sides Normal first (Section 1) and fuses via the closed-form `bayes_update_normal_normal()`
update - much cheaper per cell, at the cost of that Normal assumption.

```{r, fig.width = 7, fig.height = 4.5}
route_df <- data.frame(
  value = c(terra::values(fused_general$posterior$mu, na.rm = TRUE),
            terra::values(fused_closed_form$posterior$mu, na.rm = TRUE)),
  route = rep(c("General (KDE)", "Closed-form (Normal-Normal)"),
              c(sum(!is.na(terra::values(fused_general$posterior$mu))),
                sum(!is.na(terra::values(fused_closed_form$posterior$mu)))))
)
ggplot(route_df, aes(x = value, fill = route)) +
  geom_density(alpha = 0.5, color = NA) +
  scale_fill_viridis_d(name = NULL) +
  labs(title = "Fused posterior mean clay content: general vs. closed-form route",
       x = "Clay content (%)", y = "Density") +
  theme_minimal()
```

## 8. Caching internals

`run_stage1_fusion()`/`run_stage1_fusion_group()` disk-cache every fetch step under
`tools::R_user_dir("soilSIM", "cache")`, keyed by AOI + property/group id + depth window + kind.

`build_cache_key()` builds a deterministic, filename-safe key from an AOI's WKT geometry (hashed via
`digest::digest()`), a property/group id, a depth window, and a `kind` string:

```{r}
salinas_wkt <- "POLYGON((-121.66 36.60, -121.64 36.60, -121.64 36.62, -121.66 36.62, -121.66 36.60))"
aoi <- terra::project(terra::vect(salinas_wkt, crs = "epsg:4326"), "epsg:5070")

key <- build_cache_key(aoi, id = "clay", top_depth = 0, bottom_depth = 5, kind = "ssurgo")
key
CACHE_TTL_SECONDS / 86400  # cache max-age, in days
```

`cache_get(key, ttl_seconds)`/`cache_set(key, kind, value)` are the generic get/set pair every
fetch step above uses - a miss (absent, or older than `ttl_seconds`) returns `NULL` rather than
erroring:

```{r}
cache_get(key)  # NULL: nothing has been cached under this exact key in this session
cache_set(key, "ssurgo", value = list(hello = "world"))
cache_get(key)
```

### Why `SpatRaster`s need `terra::wrap()` before caching

`cache_set()`'s docs flag a real, sharp limitation: a `terra::SpatRaster` holds an **external
pointer** to in-memory/on-disk GDAL state, not the raster's actual values. A plain
`saveRDS()`/`readRDS()` round-trip serializes that pointer, which is meaningless in a new R
session - the read-back object *looks* like a `SpatRaster` but errors the moment anything tries to
touch its actual data:

```{r, error = TRUE}
r <- fusion_clay$posterior$mu
naive_file <- tempfile(fileext = ".rds")
saveRDS(r, naive_file)               # serializes the pointer, not the raster's values
r_naive <- readRDS(naive_file)
class(r_naive)                       # still LOOKS like a SpatRaster...
terra::values(r_naive)               # ...but touching it throws
```

`terra::wrap()` converts a `SpatRaster` into a `PackedSpatRaster` - a plain, fully self-contained R
object holding the actual cell values - which *does* survive `saveRDS()`/`readRDS()`;
`terra::unwrap()` converts it back:

```{r}
wrap_file <- tempfile(fileext = ".rds")
saveRDS(terra::wrap(r), wrap_file)
r_wrapped_read <- readRDS(wrap_file)
class(r_wrapped_read)                # PackedSpatRaster - plain data, not a live pointer

r_restored <- terra::unwrap(r_wrapped_read)
class(r_restored)                    # SpatRaster again
identical(terra::values(r), terra::values(r_restored))  # values round-tripped exactly
```

Since `cache_get()`/`cache_set()` themselves don't wrap/unwrap automatically (not every cached value
is a raster - e.g. cached SSURGO Monte Carlo draws are a plain data frame), every caller that caches
raster values has to do this manually. Two helpers cover the two result shapes this pipeline
actually caches:

- `wrap_percentile_list()`/`unwrap_percentile_list()` - for the fixed `list(values = <SpatRasters>,
  probs = )` percentile shape `fetch_ssurgo_percentiles()`/`fetch_solus_percentiles()` return.
- `wrap_nested_rasters()`/`unwrap_nested_rasters()` - a fully generic version for result structures
  whose shape isn't that fixed form, e.g. `run_stage1_fusion()`'s own return value, which nests
  `SpatRaster`s at `prior$values`, `likelihood$values`, and inside `posterior` (whose shape varies
  by `dist`). It recurses into every list element, replacing each `SpatRaster` it finds with its
  wrapped/unwrapped equivalent and leaving everything else untouched - which is exactly how this
  vignette's own `fusion_clay` object was loaded, back in Section "Reusing real data instead of
  fetching":

```{r}
# This is the call this whole vignette started with:
raw <- readRDS(system.file("extdata", "fusion_clay_salinas.rds", package = "soilSIM"))
class(raw$posterior$mu)                              # still PackedSpatRaster on disk
fusion_clay_reloaded <- unwrap_nested_rasters(raw)
class(fusion_clay_reloaded$posterior$mu)              # SpatRaster after unwrap_nested_rasters()
```

## Where this data came from

The cached data this vignette loads (`inst/extdata/fusion_clay_salinas.rds`) is the same object
`raster-fusion-ssurgo-solus.Rmd` loads, produced once by `data-raw/build_vignette_data.R` against
live NRCS Soil Data Access (SSURGO) and `soilDB::fetchSOLUS()` (SOLUS100) for the Salinas Valley
AOI. See that vignette for the full fetch-through-fuse pipeline view, and
`soilSIM/docs/09_multi_source_raster_fusion_pipeline.md` for the complete function-level reference
this vignette exposed piece by piece.
