---
title: "Statistics & Diagnostics: Descriptive Stats, Correlation, and QA"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Statistics & Diagnostics: Descriptive Stats, Correlation, and QA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6, fig.height = 4)
```

## Overview

soilSIM's other vignettes walk through the *pipeline* - calling top-level orchestrators like
`analyze_soil_statistics()` end to end on real SSURGO data. This vignette instead walks through the
*individual functions* that make up soilSIM's statistics-and-diagnostics layer, one at a time, so you
can see what each parameter actually controls:

1. Descriptive statistics - `analyze_soil_statistics()` and `compute_property_statistics()`
2. Correlation analysis - `run_comprehensive_correlation_analysis()` and
   `compute_stratified_correlations()` (and what "stratified" means)
3. Distribution shape analysis - `analyze_property_distributions()`, `fit_property_distributions()`,
   `get_appropriate_distributions()`
4. Outlier detection - `detect_comprehensive_outliers()`
5. Texture-specific correlation handling - `analyze_texture_correlations()`
6. The QA/reporting layer - `validate_statistical_results()`, `generate_statistical_quality_report()`,
   `get_statistical_analysis_defaults()`, `validate_statistical_config()`

See `soilSIM/docs/02_statistics_diagnostics.md` for the full function-level architecture reference
behind this group.

Because this vignette is about statistical *behavior* rather than any particular real dataset, it
uses small, synthetic soil-property data built with a **known** correlation structure and a handful
of **known** outliers, so the effect of each parameter can be isolated from real-data messiness. None
of the numbers below describe an actual soil - see "The synthetic dataset" for exactly how it was
built. (soilSIM's percentile-triplet distribution-fitting *core*, `R/distributions.R`, is covered in a
sibling vignette; this one stays focused on how the statistics layer *uses* distribution fitting.)

```{r setup}
library(soilSIM)
library(ggplot2)
```

## The synthetic dataset

We build 90 synthetic "horizons" split evenly across three genhz-style groups (`"A"`, `"Bt"`, `"C"`).
Clay content, organic matter, and pH are drawn from a multivariate-normal latent space with a
**known** correlation matrix (clay-om positively correlated, clay-pH and om-pH negatively
correlated), then mapped onto realistic property scales. Sand and silt are derived so that
`sandtotal_r + silttotal_r + claytotal_r == 100` for every row - a real compositional constraint that
section 5 (`analyze_texture_correlations()`) specifically deals with.

```{r}
set.seed(2026)

n <- 90
genhz <- rep(c("A", "Bt", "C"), each = n / 3)

# Known correlation structure among the latent (standardized) variables
mu <- c(clay = 0, om = 0, ph = 0)
Sigma <- matrix(
  c(1.00,  0.55, -0.45,
    0.55,  1.00, -0.30,
   -0.45, -0.30,  1.00),
  nrow = 3, byrow = TRUE, dimnames = list(names(mu), names(mu))
)
latent <- MASS::mvrnorm(n, mu = mu, Sigma = Sigma)

claytotal_r  <- pmin(pmax(25 + 8 * latent[, "clay"], 5), 60)
om_r         <- pmin(pmax(2.5 + 1.2 * latent[, "om"], 0.1), 8)
ph1to1h2o_r  <- pmin(pmax(6.3 + 0.6 * latent[, "ph"], 4.5), 8.3)

# Close the sand/silt/clay texture triangle to 100
sand_share   <- pmin(pmax(rnorm(n, 0.55, 0.1), 0.15), 0.85)
sandtotal_r  <- (100 - claytotal_r) * sand_share
silttotal_r  <- 100 - claytotal_r - sandtotal_r

soil_synth <- data.frame(
  cokey       = sprintf("synthetic-%03d", seq_len(n)),
  genhz       = genhz,
  hzdept_r    = rep(c(0, 20, 45), each = n / 3) + round(runif(n, -3, 3)),
  claytotal_r = claytotal_r,
  sandtotal_r = sandtotal_r,
  silttotal_r = silttotal_r,
  om_r        = om_r,
  ph1to1h2o_r = ph1to1h2o_r
)
```

Finally, three deliberate, **known** outliers are injected so `detect_comprehensive_outliers()` in
section 4 can be shown actually catching them, rather than just asserting that it works:

```{r}
outlier_idx <- c(5, 40, 85)
soil_synth$claytotal_r[outlier_idx[1]]  <- 95   # implausibly high clay
soil_synth$ph1to1h2o_r[outlier_idx[2]]  <- 3.0  # implausibly acidic
soil_synth$om_r[outlier_idx[3]]         <- 15   # implausibly high organic matter

soil_synth[outlier_idx, c("cokey", "claytotal_r", "om_r", "ph1to1h2o_r")]
```

A flat configuration list is used throughout sections 2-4, matching what
`run_comprehensive_correlation_analysis()`, `compute_stratified_correlations()`,
`analyze_texture_correlations()`, and `detect_comprehensive_outliers()` each read directly off their
`config` argument (`config$stratify_by_horizon`, `config$minimum_observations`, and so on):

```{r}
flat_config <- list(
  minimum_observations       = 10,
  handle_constant_variables  = "warn",
  stratify_by_horizon        = TRUE,
  include_texture_analysis   = TRUE,
  outlier_methods             = c("iqr", "zscore", "modified_zscore"),
  outlier_thresholds          = list(iqr = 1.5, zscore = 3, modified_zscore = 3.5),
  detect_multivariate_outliers = TRUE,
  mahalanobis_alpha           = 0.975
)
```

Note that this is *not* the same shape as `get_statistical_analysis_defaults()`'s return value (see
section 6) - that function nests these same keys under a `$statistical_analysis` element for
`analyze_soil_statistics()`'s own internal use, while the standalone functions demonstrated below read
them unnested.

## 1. Descriptive statistics

### `analyze_soil_statistics()` - the master entry point

`analyze_soil_statistics()` is the orchestrator used by the getting-started vignette: validate data
quality, identify numeric properties, handle missing values, then run correlation, distribution, and
outlier analysis before assembling a quality report. Two of its arguments directly toggle whole
pipeline stages - `distribution_fitting` and `outlier_detection` - so their effect is visible in the
shape of the returned list, not just in log output:

```{r}
stats_full <- analyze_soil_statistics(
  soil_synth,
  distribution_fitting = TRUE,
  outlier_detection = TRUE,
  verbose = FALSE
)
names(stats_full)
is.null(stats_full$outlier_analysis)

stats_no_outliers <- analyze_soil_statistics(
  soil_synth,
  distribution_fitting = TRUE,
  outlier_detection = FALSE,
  verbose = FALSE
)
is.null(stats_no_outliers$outlier_analysis)
```

`analysis_config` is merged on top of `get_statistical_analysis_defaults()` via
`merge_configurations()`; `correlation_methods` controls which correlation methods Step 4 computes.
Per-property descriptive statistics (from Step 7's internal `compute_property_statistics()` call - see
below) are always present:

```{r}
stats_full$property_statistics$claytotal_r[c("n_observations", "mean", "sd", "skewness", "kurtosis")]
```

### `compute_property_statistics()` - the descriptive-statistics building block

`compute_property_statistics(data, properties, config)` is what Step 7 above calls internally, and can
be called directly on any data frame. For each property it computes n/missing-rate/mean/sd/min/max/
range/CV/median/quartiles/IQR on finite values, a 95% t-based confidence interval, skewness, kurtosis,
and (for n <= 5000) a Shapiro-Wilk normality p-value.

The injected clay outlier (row `r outlier_idx[1]`, clay = 95) is a good demonstration of how sensitive
plain moment-based statistics are to a single extreme value - compare the full-data statistics against
the same computation with that one row excluded:

```{r}
prop_stats <- compute_property_statistics(
  soil_synth,
  properties = c("claytotal_r", "om_r", "ph1to1h2o_r"),
  config = list()
)
prop_stats$claytotal_r[c("mean", "sd", "skewness", "shapiro_p")]

prop_stats_no_outlier <- compute_property_statistics(
  soil_synth[-outlier_idx[1], ],
  properties = "claytotal_r",
  config = list()
)
prop_stats_no_outlier$claytotal_r[c("mean", "sd", "skewness", "shapiro_p")]
```

A single extreme value moves the mean, inflates the standard deviation, and pushes skewness and the
Shapiro-Wilk p-value toward "not normal" - exactly the symptom `detect_comprehensive_outliers()`
(section 4) is built to flag before it reaches conclusions like this one.

## 2. Correlation analysis

### `run_comprehensive_correlation_analysis()` - the full correlation routine

This is the richer, exported correlation function (as opposed to the lighter internal `_safe` version
`analyze_soil_statistics()` uses in Step 4) - it additionally computes eigenvalues and a condition
number per matrix, and can add horizon-stratified and texture-specific blocks depending on `config`:

```{r}
properties <- c("claytotal_r", "om_r", "ph1to1h2o_r")

corr_result <- run_comprehensive_correlation_analysis(
  data = soil_synth,
  methods = c("pearson", "spearman"),
  config = flat_config,
  available_properties = properties
)
names(corr_result)
corr_result$matrices$pearson$matrix
corr_result$matrices$pearson$condition_number
```

The recovered Pearson correlations (~0.5-0.6 for clay-om, ~-0.4 for clay-pH, ~-0.3 for om-pH) track
the `Sigma` matrix the synthetic data was built from in the previous section - correlation analysis is
being validated against a *known* ground truth here, not just checked for "doesn't error."

```{r, fig.height = 4}
cor_df <- as.data.frame(as.table(corr_result$matrices$pearson$matrix))
names(cor_df) <- c("property_1", "property_2", "correlation")

ggplot(cor_df, aes(x = property_1, y = property_2, fill = correlation)) +
  geom_tile() +
  geom_text(aes(label = sprintf("%.2f", correlation)), color = "white", size = 3.5) +
  scale_fill_viridis_c(limits = c(-1, 1), name = "Correlation") +
  theme_minimal() +
  labs(title = "Pearson correlation, synthetic properties", x = NULL, y = NULL)
```

`config$include_texture_analysis` and `config$stratify_by_horizon` (both `TRUE` in `flat_config`)
turned on two extra blocks - `texture_analysis` (section 5) and `stratified` (below), computed only
when a `genhz` column exists:

```{r}
names(corr_result$texture_analysis)
names(corr_result$stratified)
```

### `compute_stratified_correlations()` - what "stratified" means here

A single flat correlation matrix pools all 90 synthetic horizons together, across all three genhz
groups. But if a variable (here, depth-driven `genhz`) itself has different property means in each
group, a pooled correlation can be a **weighted blend of the group-level correlations plus a
between-group component driven purely by group membership** - not necessarily what any single group
actually shows. `compute_stratified_correlations(data, properties, stratify_by, methods, config)`
splits the data by the unique values of `stratify_by` (here `"genhz"`), skips any group with fewer
than `config$minimum_observations` rows, and computes one correlation matrix per method *within* each
remaining group:

```{r}
stratified <- compute_stratified_correlations(
  data = soil_synth,
  properties = properties,
  stratify_by = "genhz",
  methods = "pearson",
  config = flat_config
)
names(stratified)
stratified$A$pearson$matrix
stratified$A$pearson$n_observations
```

Compare one group's matrix against the pooled matrix from `run_comprehensive_correlation_analysis()`
above - because the synthetic clay/om/pH latent draws here were generated with the *same* correlation
structure inside every genhz group (no group-level confounding was built in on purpose), the two
should look similar; in real SSURGO data, where horizon genesis drives systematic depth trends, a
noticeably different stratified-vs-pooled result is the signal that stratification matters.

`config$minimum_observations` directly controls which groups survive. With 30 rows per group, raising
it above 30 drops every group from the output:

```{r}
stratified_strict <- compute_stratified_correlations(
  data = soil_synth,
  properties = properties,
  stratify_by = "genhz",
  methods = "pearson",
  config = modifyList(flat_config, list(minimum_observations = 50))
)
length(stratified_strict)
```

## 3. Distribution shape analysis

### `get_appropriate_distributions()` - candidate family selection

`get_appropriate_distributions(property_name, values, config)` picks candidate distribution families
using a type-based heuristic keyed off the property *name*: `c("beta", "normal")` for the three
texture fractions, `c("normal", "gamma")` for `ph1to1h2o_r`, and `c("normal", "lognormal", "gamma",
"weibull")` for everything else. `values` is currently unused by the heuristic itself (kept for
interface symmetry).

```{r}
get_appropriate_distributions("claytotal_r", soil_synth$claytotal_r)
get_appropriate_distributions("ph1to1h2o_r", soil_synth$ph1to1h2o_r)
get_appropriate_distributions("om_r", soil_synth$om_r)
```

`config$distribution_methods`, when supplied, narrows the heuristic to its intersection with the
requested list - a soft prior rather than a silent override:

```{r}
get_appropriate_distributions("claytotal_r", soil_synth$claytotal_r,
                               config = list(distribution_methods = c("normal", "gamma")))
```

If the intersection is empty (asking for families the heuristic wouldn't have suggested for this
property), the function falls back to the requested list unfiltered rather than discarding it, with a
logged warning:

```{r}
get_appropriate_distributions("claytotal_r", soil_synth$claytotal_r,
                               config = list(distribution_methods = "weibull"))
```

### `fit_property_distributions()` - fitting and ranking candidates

`fit_property_distributions(values, property_name, config)` fits every candidate from
`get_appropriate_distributions()` via `fitdistrplus::fitdist()`, skips any that fail to converge, and
ranks the survivors best-first by AIC:

```{r}
om_fits <- fit_property_distributions(soil_synth$om_r, "om_r", config = list())
names(om_fits)
sapply(om_fits, function(f) f$aic)
```

The best-fitting family (lowest AIC, listed first) here should recover something close to
`om_r`'s true generative shape - a normal distribution truncated/clamped to `[0.1, 8]` - though a
`lognormal` or `gamma` fit can sometimes edge it out slightly given the clamping and finite sample
size.

### `analyze_property_distributions()` - the full per-property pipeline

`analyze_property_distributions(data, properties, config)` combines fitting with goodness-of-fit
testing (`perform_distribution_tests()`, via `fitdistrplus::gofstat()`) for every property with at
least `config$minimum_observations` finite values:

```{r}
dist_result <- analyze_property_distributions(
  soil_synth,
  properties = c("om_r", "ph1to1h2o_r"),
  config = flat_config
)
names(dist_result$fitted_distributions$ph1to1h2o_r)
dist_result$summary
```

This is distinct from the *statistics-layer* usage above only in scope - the underlying
percentile-triplet distribution-fitting machinery that turns each SSURGO horizon's low/representative/
high values into a single fitted distribution for Monte Carlo simulation lives in `R/distributions.R`
and is covered in its own sibling vignette.

## 4. Outlier detection

### `detect_comprehensive_outliers()` - univariate and multivariate

`detect_comprehensive_outliers(data, properties, config)` runs every method in
`config$outlier_methods` per property (using the matching threshold from
`config$outlier_thresholds`), and - when `config$detect_multivariate_outliers` is `TRUE` and at least
two properties are supplied - a real Mahalanobis-distance multivariate check flagging the outer
`1 - config$mahalanobis_alpha` fraction via a chi-squared cutoff.

```{r}
outlier_result <- detect_comprehensive_outliers(
  data = soil_synth,
  properties = properties,
  config = flat_config
)
names(outlier_result$property_outliers$claytotal_r)
outlier_result$property_outliers$claytotal_r$iqr$n_outliers
```

Do the univariate methods actually catch the three injected outliers at their known row positions
(`r paste(outlier_idx, collapse = ", ")`)? The IQR method's `outliers` vector is indexed against the
finite values of that property in row order, so for `claytotal_r` (no missing values) row 5's flag can
be read directly:

```{r}
outlier_result$property_outliers$claytotal_r$iqr$outliers[outlier_idx[1]]
outlier_result$property_outliers$ph1to1h2o_r$iqr$outliers[outlier_idx[2]]
outlier_result$property_outliers$om_r$iqr$outliers[outlier_idx[3]]
```

Each injected outlier is caught by the IQR check on its own property. The multivariate check looks
across all three properties jointly via Mahalanobis distance:

```{r}
outlier_result$multivariate_outliers$n_outliers
which(outlier_result$multivariate_outliers$outliers)
```

`config$outlier_thresholds$iqr` controls how aggressive the IQR method is (multiplier on the IQR
beyond Q1/Q3) - the conventional default is 1.5; doubling it should flag fewer, more extreme points:

```{r}
loose_config <- modifyList(flat_config, list(outlier_thresholds = list(iqr = 3, zscore = 3, modified_zscore = 3.5)))
outlier_loose <- detect_comprehensive_outliers(soil_synth, properties, loose_config)
sum(outlier_loose$property_outliers$claytotal_r$iqr$outliers)
sum(outlier_result$property_outliers$claytotal_r$iqr$outliers)
```

`config$mahalanobis_alpha` controls the multivariate flagging rate directly - it is the chi-squared
percentile used as the cutoff, so a lower value flags a larger fraction of points as multivariate
outliers:

```{r}
outlier_sensitive <- detect_comprehensive_outliers(
  soil_synth, properties, modifyList(flat_config, list(mahalanobis_alpha = 0.90))
)
outlier_sensitive$multivariate_outliers$n_outliers
outlier_result$multivariate_outliers$n_outliers
```

## 5. Texture correlations: `analyze_texture_correlations()`

Sand, silt, and clay are compositional data - they are constrained to sum to 100 for every horizon, by
construction in this synthetic dataset (and, approximately, in real SSURGO data too). That constraint
alone forces some pairwise correlations to be spuriously negative: if clay goes up, *something else*
in a fixed-sum triplet has to go down, whether or not the two properties have any real relationship.
`analyze_texture_correlations()` reports the raw correlations for continuity, but also computes the
same correlations in isometric-log-ratio (ILR) space - a transform that removes the sum-to-100
constraint - which is the statistically defensible view for compositional data:

```{r}
texture_result <- analyze_texture_correlations(
  data = soil_synth,
  texture_properties = c("sandtotal_r", "silttotal_r", "claytotal_r"),
  methods = "pearson",
  config = list()
)
texture_result$raw_correlations$pearson
texture_result$ilr_correlations$pearson
texture_result$note
```

Sand and clay in particular tend to look more strongly (and more spuriously) anti-correlated in raw
space than in ILR space - the raw view is dominated by the closure constraint, while the ILR view
reflects the underlying relationship between the properties. `analyze_texture_correlations()` only
computes `ilr_correlations` when all three fractions are present with more than 4 complete rows;
otherwise it stays `NULL` and only `raw_correlations` is available.

## 6. The QA/reporting layer

### `get_statistical_analysis_defaults()` and `validate_statistical_config()`

`get_statistical_analysis_defaults()` takes `get_default_configuration("full")` and merges in a
`statistical_analysis` block covering stratification flags, missing-value strategy, correlation
defaults, distribution-fitting families, outlier defaults, and quality thresholds:

```{r}
default_config <- get_statistical_analysis_defaults()
names(default_config$statistical_analysis)
default_config$statistical_analysis$outlier_thresholds
```

`validate_statistical_config(config)` checks a statistical-analysis config against a fixed parameter
specification (types, allowed choices, numeric ranges) using `strict_mode = FALSE`, so most violations
become warnings rather than hard errors. It transparently unwraps a nested `$statistical_analysis`
block, so it accepts `get_statistical_analysis_defaults()`'s output directly:

```{r}
validate_statistical_config(default_config)$valid
```

A config with an out-of-range value (`minimum_observations` must be in `[3, 1000]`) still comes back
`valid` under non-strict mode, but with a warning identifying exactly which field failed:

```{r}
bad_config <- list(
  minimum_observations = 2000,
  missing_data_threshold = 0.8,
  min_quality_score = 0.6,
  correlation_methods = c("pearson", "spearman"),
  missing_value_strategy = "interpolate",
  error_recovery_action = "warn"
)
bad_check <- validate_statistical_config(bad_config)
bad_check$valid
bad_check$warnings
```

### `validate_statistical_results()` and `generate_statistical_quality_report()`

`validate_statistical_results()` takes the outputs assembled in the sections above and rolls them into
one validity assessment - real per-matrix checks for correlation (via `validate_correlation_matrix()`
in `R/distributions.R`), plus a validation score penalizing errors and warnings:

```{r}
validation_result <- validate_statistical_results(
  correlation_analysis = corr_result,
  distribution_analysis = dist_result,
  outlier_analysis = outlier_result,
  property_statistics = prop_stats,
  config = flat_config
)
validation_result[c("overall_valid", "validation_score")]
validation_result$warnings
```

`generate_statistical_quality_report()` composes everything above - data quality, correlation quality,
distribution quality, outlier quality, and the validation results - into a single overall quality
score plus textual recommendations. It also needs a `data_validation` object (from
`validate_data_quality()`) and both an `original_data` and `processed_data` frame - here the same
synthetic data serves as both, since this vignette does no separate infilling step:

```{r}
data_validation <- validate_data_quality(
  soil_synth,
  required_columns = c("cokey", "hzdept_r"),
  numeric_columns = identify_numeric_soil_properties(soil_synth)
)

quality_report <- generate_statistical_quality_report(
  original_data = soil_synth,
  processed_data = soil_synth,
  correlation_analysis = corr_result,
  distribution_analysis = dist_result,
  outlier_analysis = outlier_result,
  property_statistics = prop_stats,
  validation_results = validation_result,
  data_validation = data_validation,
  config = flat_config
)
quality_report$overall_quality_score
quality_report$analysis_quality
quality_report$recommendations
```

This is the same composite report structure `analyze_soil_statistics()` returns as
`$quality_report` in section 1 - `generate_statistical_quality_report()` is simply that step made
callable on its own, over whatever correlation/distribution/outlier results you have already computed
by hand.
